---
title: "Gene Ontology & Pathway Enrichment Analysis with g:Profiler2"
author: "Phuong Nguyen"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 4
    number_sections: true
    code_folding: hide
    df_print: paged
    fig_width: 10
    fig_height: 7
---

```{r setup, include=FALSE}
################################################################################
# Global knitr options
# - echo = TRUE     : show code by default (user can toggle via code_folding)
# - warning/message : suppressed for cleaner report output
# - fig.align       : center all figures
# - dpi             : 150 for crisp HTML rendering; increase to 300 for print
################################################################################
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.align  = "center",
  dpi        = 150,
  out.width  = "100%"
)
```

# Introduction

This pipeline performs **functional enrichment analysis** using the
[g:Profiler2](https://biit.cs.ut.ee/gprofiler/gost) R client. g:Profiler
queries curated databases including Gene Ontology (BP, CC, MF), KEGG,
Reactome, WikiPathways, TRANSFAC, miRTarBase, CORUM, Human Protein Atlas,
and HP (Human Phenotype Ontology).

**Workflow overview:**

1. Define input gene list (Ensembl IDs or gene symbols) and select organism.
2. Run enrichment via `gprofiler2::gost()` with customisable parameters.
3. Export full results as TSV/CSV tables.
4. Visualise results with g:Profiler's built-in interactive Manhattan plot
   (`gostplot`) and enrichment table (`publish_gosttable`).
5. Generate publication-quality ggplot2 bar charts, dot plots, and lollipop
   plots — faceted by ontology source — styled after the reference figure.

# Package Installation & Loading

```{r install-packages, eval=FALSE}
################################################################################
# Run this chunk ONCE to install required packages. Set eval=TRUE if needed.
# gprofiler2 : R client for g:Profiler web service
# ggplot2    : grammar-of-graphics plotting
# dplyr/tidyr: data wrangling
# DT         : interactive HTML tables
# scales     : axis formatting helpers
# forcats    : factor manipulation for ggplot ordering
# stringr    : string manipulation (term name wrapping)
# plotly     : interactive ggplot conversion (optional)
# RColorBrewer: colour palettes
################################################################################
install.packages(c(
  "gprofiler2", "ggplot2", "dplyr", "tidyr",
  "DT", "scales", "forcats", "stringr",
  "plotly", "RColorBrewer", "ggtext", "tidytext"
))
```

```{r load-packages}
################################################################################
# Load all required libraries
################################################################################
library(gprofiler2)
library(ggplot2)
library(dplyr)
library(tidyr)
library(DT)
library(scales)
library(forcats)
library(stringr)
library(plotly)
library(RColorBrewer)
library(ggtext)
library(tidytext)
```

# User-Defined Parameters {.tabset}

Edit the parameters below to customise the analysis for your dataset.

## Gene List & Organism

```{r user-parameters}
################################################################################
# ========================  EDIT THIS SECTION  ================================
#
# GENE INPUT FILE
# ---------------
# gene_file : path to a file containing gene identifiers.
#
#   Supported formats (auto-detected by file extension):
#
#   1) Plain text (.txt)
#      One gene per line. Blank lines and lines starting with "#" are ignored.
#      Example:
#        TP53
#        BRCA1
#        ENSG00000141510
#
#   2) Tab-separated (.tsv) or comma-separated (.csv)
#      The FIRST COLUMN is taken as gene IDs.
#      A header row is expected (set gene_file_header = FALSE below if absent).
#      Example TSV:
#        GeneID    log2FC    padj
#        TP53      2.31      0.0004
#        BRCA1     -1.87     0.012
#
#   Gene IDs can be Ensembl IDs (e.g. "ENSG00000141510") or gene symbols
#   (e.g. "TP53"). Mixed types are allowed — g:Profiler auto-detects the
#   namespace.
#
# gene_file_header : does the TSV/CSV file have a header row?
#   TRUE  (default) — first row is column names, skip it
#   FALSE — no header; first row is data
#   (ignored for .txt files)
#
# ORGANISM
# --------
# organism : g:Profiler organism code. Common choices:
#   "hsapiens"     — Homo sapiens
#   "mmusculus"    — Mus musculus
#   "rnorvegicus"  — Rattus norvegicus
#   "drerio"       — Danio rerio (zebrafish)
#   "ggallus"      — Gallus gallus (chicken)
#   "sscrofa"      — Sus scrofa (pig)
#   "olatipes"     — Oryzias latipes (medaka)
#   "icpunctatus"  — Ictalurus punctatus (channel catfish)
#   "oniloticus"   — Oreochromis niloticus (tilapia)
#   Full list: https://biit.cs.ut.ee/gprofiler/page/organism-list
#
# BACKGROUND (optional)
# ---------------------
# bg_file : path to a background gene list file (same format rules as above).
#           Set to NULL to use the full annotated genome (default).
#
# ========================  EDIT THIS SECTION  ================================
################################################################################

gene_file        <- "sample_input_Ggallus.txt"   # <-- CHANGE to your file path
gene_file_header <- FALSE                 # TRUE if TSV/CSV has a header row

organism         <- "ggallus"           # <-- CHANGE to your organism

bg_file          <- NULL                 # NULL = whole genome background
# bg_file        <- "background_genes.txt"   # uncomment & set path if needed
```

```{r read-gene-file}
################################################################################
# Helper function: read genes from a file.
#
# Logic:
#   - .txt  → readLines(), strip blanks & comments, trim whitespace
#   - .tsv  → read.delim(), extract first column
#   - .csv  → read.csv(),   extract first column
#   - other → attempt tab-separated read as fallback
#
# Returns a character vector of unique, non-empty gene identifiers.
################################################################################

read_gene_list <- function(filepath, has_header = TRUE) {

  if (!file.exists(filepath)) {
    stop("Gene file not found: ", filepath,
         "\n  Please check the path in the 'gene_file' parameter.")
  }

  ext <- tolower(tools::file_ext(filepath))

  if (ext == "txt") {
    ## --- Plain text: one gene per line ---
    raw <- readLines(filepath, warn = FALSE)
    ## Remove comment lines (starting with #) and blank lines
    raw <- raw[!grepl("^\\s*#", raw)]
    raw <- trimws(raw)
    genes <- raw[nchar(raw) > 0]

  } else if (ext %in% c("tsv", "csv")) {
    ## --- Tabular: take the first column ---
    sep <- ifelse(ext == "csv", ",", "\t")
    df  <- read.delim(filepath, sep = sep, header = has_header,
                      stringsAsFactors = FALSE, check.names = FALSE,
                      quote = "\"", comment.char = "#")
    genes <- trimws(as.character(df[[1]]))
    genes <- genes[nchar(genes) > 0 & !is.na(genes)]

  } else {
    ## --- Fallback: try tab-separated ---
    message("Unrecognised extension '.", ext,
            "' — attempting tab-separated read.")
    df  <- read.delim(filepath, sep = "\t", header = has_header,
                      stringsAsFactors = FALSE, check.names = FALSE)
    genes <- trimws(as.character(df[[1]]))
    genes <- genes[nchar(genes) > 0 & !is.na(genes)]
  }

  ## Remove duplicates and return
  genes <- unique(genes)

  if (length(genes) == 0) {
    stop("No gene identifiers found in: ", filepath,
         "\n  Check file content and format.")
  }

  return(genes)
}

## ---- Read the query gene list ----
gene_list <- read_gene_list(gene_file, has_header = gene_file_header)

cat("Gene list file :", gene_file, "\n")
cat("Genes loaded   :", length(gene_list), "\n")
cat("First 10 genes :", paste(head(gene_list, 10), collapse = ", "), "\n")

## ---- Read background gene list (if provided) ----
if (!is.null(bg_file)) {
  background_genes <- read_gene_list(bg_file, has_header = gene_file_header)
  cat("\nBackground file :", bg_file, "\n")
  cat("Background genes:", length(background_genes), "\n")
} else {
  background_genes <- NULL
  cat("\nBackground: full annotated genome (default)\n")
}
```

## Generate Demo Gene List (Optional) {.unlisted .unnumbered}

Run the chunk below **once** to create example input files for testing.
Skip this if you already have your own gene list file.

```{r create-demo-files, eval=FALSE, include=FALSE}
################################################################################
# Creates two demo files in the working directory:
#   1) my_gene_list.txt — plain text, one gene symbol per line
#   2) my_gene_list.tsv — tab-separated with GeneID, log2FC, padj columns
#
# Set eval=TRUE (or run manually) to generate these files, then knit the
# full report.
################################################################################

demo_genes <- c(
  "TP53", "BRCA1", "BRCA2", "EGFR", "VEGFA", "MYC", "KRAS", "PTEN",
  "AKT1", "PIK3CA", "MTOR", "CDK4", "CDK6", "RB1", "CCND1", "CCNE1",
  "BCL2", "BAX", "CASP3", "CASP8", "CASP9", "APAF1", "CYCS", "BID",
  "XIAP", "BIRC5", "MCL1", "BCL2L1", "FAS", "FASLG", "TNFRSF10A",
  "TNFRSF10B", "TRAIL", "TRAF2", "RIPK1", "MLKL", "ATM", "ATR",
  "CHEK1", "CHEK2", "MDM2", "CDKN1A", "CDKN2A", "GADD45A", "E2F1",
  "DIABLO", "HTRA2", "ENDOG", "AIFM1", "DFFB"
)

## Plain text version
writeLines(demo_genes, "my_gene_list.txt")
cat("Created: my_gene_list.txt\n")

## TSV version (with dummy fold-change & p-value columns)
set.seed(42)
demo_df <- data.frame(
  GeneID = demo_genes,
  log2FC = round(rnorm(length(demo_genes), mean = 0, sd = 2), 3),
  padj   = signif(runif(length(demo_genes), 1e-6, 0.05), 3),
  stringsAsFactors = FALSE
)
write.table(demo_df, "my_gene_list.tsv", sep = "\t",
            row.names = FALSE, quote = FALSE)
cat("Created: my_gene_list.tsv\n")
```

## Advanced g:Profiler Parameters

```{r advanced-parameters}
################################################################################
# Advanced parameters passed to gost().
# Most users can leave these at their defaults.
#
# correction_method : multiple-testing correction. Options:
#     "g_SCS" (default, g:Profiler's own method),
#     "fdr"   (Benjamini-Hochberg),
#     "bonferroni"
#
# user_threshold    : adjusted p-value cutoff for significance.
#
# sources           : annotation databases to query.
#     NULL (default) queries all available sources.
#     Restrict with e.g. c("GO:BP", "GO:CC", "GO:MF", "KEGG", "REAC", "WP")
#
# significant       : if TRUE, return only terms passing the threshold;
#                     if FALSE, return all tested terms.
#
# evcodes           : if TRUE, include evidence codes and intersection gene
#                     lists (slower but informative).
#
# top_n             : number of top terms per source to show in plots.
################################################################################

correction_method <- "g_SCS"
user_threshold    <- 0.05
sources           <- NULL           # NULL = all sources
significant       <- TRUE
evcodes           <- TRUE           # include intersection genes
top_n             <- 10             # top N terms per source for plots
```


# Run g:Profiler Enrichment Analysis

The core function `gost()` sends the gene list to the g:Profiler server
and returns enrichment results across all selected databases.

```{r run-gost}
################################################################################
# Run the enrichment query.
# Key arguments:
#   query              — input gene list
#   organism           — target species
#   ordered_query      — FALSE: unordered list (standard ORA)
#   multi_query        — FALSE: single gene list
#   significant        — filter by significance
#   exclude_iea        — FALSE: include electronically-inferred GO annotations
#   measure_underrepresentation — FALSE: test for over-representation only
#   evcodes            — include evidence codes & intersection genes
#   user_threshold     — significance cutoff
#   correction_method  — multiple-testing method
#   domain_scope       — "annotated": use annotated genes as background
#   custom_bg          — user-defined background (NULL = full genome)
#   sources            — databases to query
################################################################################

gost_result <- gost(
  query             = gene_list,
  organism          = organism,
  ordered_query     = FALSE,
  multi_query       = FALSE,
  significant       = significant,
  exclude_iea       = FALSE,
  measure_underrepresentation = FALSE,
  evcodes           = evcodes,
  user_threshold    = user_threshold,
  correction_method = correction_method,
  ## domain_scope must be "custom" when a background gene list is supplied;
  ## otherwise "annotated" uses all genes with annotations in that organism.
  domain_scope      = ifelse(is.null(background_genes), "annotated", "custom"),
  custom_bg         = background_genes,
  sources           = sources
)
```

```{r check-results}
################################################################################
# Verify that the query returned results before proceeding.
################################################################################
if (is.null(gost_result) || is.null(gost_result$result) || nrow(gost_result$result) == 0) {
  stop("No significant enrichment terms found. ",
       "Try relaxing the threshold, expanding the gene list, ",
       "or checking gene ID format / organism selection.")
}

cat("Total significant terms found:", nrow(gost_result$result), "\n")
cat("Sources represented:\n")
print(table(gost_result$result$source))
```

# Enrichment Result Table

## Full Result Overview

```{r result-overview}
################################################################################
# gost() returns a list with two elements:
#   $result — data.frame of enrichment results
#   $meta   — metadata (organism, version, timestamp, query mapping)
#
# Key columns in $result:
#   source          — database (GO:BP, GO:CC, GO:MF, KEGG, REAC, WP, ...)
#   term_id         — accession (e.g. GO:0006915, KEGG:04210)
#   term_name       — human-readable term description
#   p_value         — adjusted p-value (after correction_method)
#   term_size       — total genes annotated to this term
#   query_size      — number of query genes with any annotation
#   intersection_size — overlap: query genes in this term
#   precision       — intersection_size / query_size
#   recall          — intersection_size / term_size
#   intersection    — comma-separated list of overlapping genes (if evcodes=T)
################################################################################

## Create a clean results data.frame for downstream use
enrich_df <- gost_result$result %>%
  dplyr::select(
    source, term_id, term_name, p_value,
    term_size, query_size, intersection_size,
    precision, recall,
    tidyselect::any_of(c("intersection", "effective_domain_size"))
  ) %>%
  dplyr::arrange(source, p_value)

## Compute additional metrics useful for plotting
enrich_df <- enrich_df %>%
  dplyr::mutate(
    ## Gene ratio: fraction of query genes in this term
    gene_ratio  = intersection_size / query_size,
    ## Fold enrichment: observed / expected overlap
    ## NOTE: effective_domain_size varies by source (GO, KEGG, Reactome each
    ## have different numbers of annotated genes), so it must be used per-row.
    fold_enrich = (intersection_size / query_size) /
                  (term_size / effective_domain_size),
    ## -log10(p-value) for volcano-style visualisation
    neg_log10_p = -log10(p_value),
    ## log10(fold enrichment) — matches the reference figure x-axis
    log10_fe    = log10(pmax(fold_enrich, 1))  # floor at 1 to avoid negative log
  )

cat("Result table dimensions:", nrow(enrich_df), "rows x", ncol(enrich_df), "columns\n")
```

## Interactive Table (Top Results)

```{r interactive-table}
################################################################################
# Display the top enrichment results per source as an interactive DT table.
# Users can search, sort, and page through results in the HTML report.
################################################################################
top_table <- enrich_df %>%
  dplyr::group_by(source) %>%
  dplyr::slice_min(p_value, n = top_n, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::select(source, term_id, term_name, p_value,
                intersection_size, term_size, fold_enrich, gene_ratio)

DT::datatable(
  top_table,
  caption  = paste0("Top ", top_n, " enriched terms per source (sorted by p-value)"),
  filter   = "top",
  rownames = FALSE,
  options  = list(
    pageLength = 20,
    scrollX    = TRUE,
    columnDefs = list(
      list(className = "dt-center", targets = "_all")
    )
  )
) %>%
  DT::formatSignif(columns = c("p_value", "fold_enrich", "gene_ratio"), digits = 4)
```

# Export Results to File

```{r export-tables}
################################################################################
# Export the full enrichment results as both TSV and CSV files.
# These files are saved in the working directory.
# The 'intersection' column (if present) lists the overlapping genes per term.
################################################################################

## Prepare export data.frame (remove plotting helper columns)
export_df <- enrich_df %>%
  dplyr::select(-neg_log10_p, -log10_fe)

## TSV export (tab-separated, preferred for bioinformatics tools)
tsv_file <- paste0("gprofiler2_enrichment_", organism, "_full_results.tsv")
write.table(export_df, file = tsv_file, sep = "\t",
            row.names = FALSE, quote = FALSE)
cat("Exported TSV:", tsv_file, "\n")

## CSV export (comma-separated, for Excel / spreadsheets)
csv_file <- paste0("gprofiler2_enrichment_", organism, "_full_results.csv")
write.csv(export_df, file = csv_file, row.names = FALSE)
cat("Exported CSV:", csv_file, "\n")

## Additionally export per-source tables for convenience
for (src in unique(export_df$source)) {
  src_df   <- dplyr::filter(export_df, source == src)
  src_file <- paste0("gprofiler2_enrichment_", organism, "_", src, ".tsv")
  src_file <- gsub(":", "_", src_file)  # sanitise colons in source names
  write.table(src_df, file = src_file, sep = "\t",
              row.names = FALSE, quote = FALSE)
}
cat("Per-source TSV files exported.\n")
```


# g:Profiler Built-in Visualisations {.tabset}

## Interactive Manhattan Plot (gostplot)

The `gostplot()` function produces an interactive plotly Manhattan plot
where the x-axis arranges terms by source and the y-axis shows
-log10(adjusted p-value). Hover over points for details.

```{r gostplot-interactive, fig.height=6}
################################################################################
# gostplot() — g:Profiler's signature Manhattan-style visualisation.
# interactive = TRUE  → returns a plotly object (zoomable, hoverable)
# interactive = FALSE → returns a ggplot object (static, publication-ready)
################################################################################

## Interactive version (HTML-friendly)
p_manhattan <- gostplot(gost_result, capped = FALSE, interactive = TRUE)
p_manhattan
```

## Static Manhattan Plot (for publication)

```{r gostplot-static, fig.height=6, fig.width=12}
################################################################################
# Static gostplot — suitable for PDF / journal figures.
# publish_gostplot() can overlay term labels on the static plot.
################################################################################

p_static <- gostplot(gost_result, capped = FALSE, interactive = FALSE)

## Select top terms to highlight (by lowest p-value across all sources)
highlight_terms <- gost_result$result %>%
  dplyr::group_by(source) %>%
  dplyr::slice_min(p_value, n = 3, with_ties = FALSE) %>%
  dplyr::pull(term_id)

## Add labels to the static plot
p_labelled <- publish_gostplot(
  p_static,
  highlight = list(query_1 = highlight_terms)
)
print(p_labelled)
```

## Enrichment Table (publish_gosttable)

```{r gosttable, fig.height=8, fig.width=14}
################################################################################
# publish_gosttable() — creates a publication-ready table rendered as a plot.
# Includes term names, p-values, and coloured source indicators.
################################################################################

publish_gosttable(
  gost_result,
  highlight_terms = gost_result$result %>%
    dplyr::group_by(source) %>%
    dplyr::slice_min(p_value, n = 5, with_ties = FALSE) %>%
    dplyr::pull(term_id),
  use_colors = TRUE,
  show_columns = c("source", "term_name", "term_size",
                    "intersection_size"),
  filename = NULL  # NULL = display in R; set path to save as PNG
)
```


# Publication-Quality ggplot2 Visualisations {.tabset}

This section generates several types of plots styled after the uploaded
reference figure — horizontal bar charts faceted by ontology source, with
bars coloured by p-value and the x-axis showing log10(fold enrichment).

```{r plot-theme}
################################################################################
# Define a shared publication theme for all ggplot2 figures.
# This ensures consistent styling across bar, dot, and lollipop plots.
################################################################################

theme_publication <- function(base_size = 12) {
  theme_bw(base_size = base_size) %+replace%
    theme(
      ## Panel
      panel.grid.major.y = element_blank(),
      panel.grid.minor   = element_blank(),
      panel.border       = element_rect(colour = "black", fill = NA, linewidth = 0.8),
      ## Strip (facet labels)
      strip.background   = element_rect(fill = "grey90", colour = "black",
                                         linewidth = 0.5),
      strip.text         = element_text(face = "bold", size = rel(1.1)),
      ## Axis
      axis.text.y        = element_text(size = rel(0.95), colour = "grey20"),
      axis.text.x        = element_text(size = rel(0.9)),
      axis.title         = element_text(size = rel(1.0)),
      ## Legend
      legend.position    = "right",
      legend.title       = element_text(size = rel(0.9), face = "bold"),
      legend.text        = element_text(size = rel(0.8)),
      legend.key.height  = unit(1, "cm"),
      ## Plot
      plot.title         = element_text(face = "bold", hjust = 0.5, size = rel(1.2)),
      plot.subtitle      = element_text(hjust = 0.5, size = rel(0.9),
                                         colour = "grey40"),
      plot.margin        = margin(10, 15, 10, 10)
    )
}
```

```{r prepare-plot-data}
################################################################################
# Prepare per-source top-N data for plotting.
# We create a "source_label" column with cleaner names for facet strips,
# and wrap long term names for readability on the y-axis.
################################################################################

## Readable source labels
source_labels <- c(
  "GO:BP"  = "Biological Process (BP)",
  "GO:CC"  = "Cellular Component (CC)",
  "GO:MF"  = "Molecular Function (MF)",
  "KEGG"   = "KEGG Pathway",
  "REAC"   = "Reactome",
  "WP"     = "WikiPathways",
  "TF"     = "TRANSFAC (TF)",
  "MIRNA"  = "miRTarBase",
  "CORUM"  = "CORUM Complex",
  "HPA"    = "Human Protein Atlas",
  "HP"     = "Human Phenotype"
)

plot_df <- enrich_df %>%
  dplyr::group_by(source) %>%
  dplyr::slice_min(p_value, n = top_n, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    ## Clean source labels for facets
    source_label = ifelse(source %in% names(source_labels),
                          source_labels[source], source),
    source_label = factor(source_label,
                          levels = source_labels[source_labels %in% unique(source_label)]),
    ## Wrap long term names at 45 characters for y-axis readability
    term_wrap = stringr::str_wrap(term_name, width = 45)
  )

## List which sources actually have results
available_sources <- unique(plot_df$source)
cat("Sources with enrichment results:", paste(available_sources, collapse = ", "), "\n")
```

## GO Terms — Bar Chart (Reference Style) {.tabset}

### BP + CC + MF Combined

This plot replicates the style of the uploaded reference figure:
horizontal bars coloured by p-value, faceted by GO sub-ontology,
with log10(fold enrichment) on the x-axis.

```{r go-bar-combined, fig.height=6, fig.width=11}
################################################################################
# Combined GO bar chart — BP, CC, MF in one figure with free y-axis scales.
# Matches the reference image: horizontal bars, p-value colour gradient,
# log10(Fold_Enrichment) on x-axis, faceted by ontology.
################################################################################

go_sources <- c("GO:BP", "GO:CC", "GO:MF")
go_df <- plot_df %>% dplyr::filter(source %in% go_sources)

if (nrow(go_df) > 0) {
  ## Order terms within each source by fold enrichment
  go_df <- go_df %>%
    dplyr::mutate(
      term_wrap = tidytext::reorder_within(term_wrap, log10_fe, source_label)
    )

  p_go_bar <- ggplot(go_df, aes(x = log10_fe, y = term_wrap, fill = p_value)) +
    geom_col(width = 0.75, colour = "grey30", linewidth = 0.2) +
    ## Colour gradient: darker = more significant (lower p-value)
    scale_fill_gradient(
      low    = "#08306B",   # dark blue — highly significant
      high   = "#9ECAE1",   # light blue — less significant
      name   = "p-value",
      breaks = pretty_breaks(n = 4),
      guide  = guide_colorbar(reverse = FALSE)
    ) +
    ## Clean up reorder_within formatting on y-axis
    tidytext::scale_y_reordered() +
    ## Free y-axis: each facet shows its own terms
    facet_grid(source_label ~ ., scales = "free_y", space = "free_y") +
    labs(
      title    = "Selected GO Terms",
      x        = expression(log[10](Fold~Enrichment)),
      y        = NULL
    ) +
    theme_publication() +
    theme(
      strip.text.y.right  = ggtext::element_textbox(size=9,face= "bold",orientation = "upright",width= unit(2, "cm"),halign= 0.5,valign= 0.5,margin= margin(1, 1, 1,1)),
      axis.text.y        = element_text(size = 10, hjust = 1),
      plot.title = element_text(margin = margin(b = 8))
    )

  print(p_go_bar)
} else {
  cat("No GO terms found in the enrichment results.\n")
}
```

### Alternative: Without tidytext dependency

If `tidytext` is not installed, use this version with manual ordering.

```{r go-bar-alt, fig.height=6, fig.width=11}
################################################################################
# Alternative bar chart that avoids the tidytext dependency.
# Uses a combined factor level to produce correct per-facet ordering.
################################################################################

go_df2 <- plot_df %>%
  dplyr::filter(source %in% c("GO:BP", "GO:CC", "GO:MF")) %>%
  dplyr::arrange(source_label, log10_fe) %>%
  dplyr::mutate(
    ## Create unique ordered factor: paste source + term to guarantee uniqueness
    term_ordered = factor(paste(source, term_wrap, sep = "___"),
                          levels = paste(source, term_wrap, sep = "___"))
  )

if (nrow(go_df2) > 0) {
  p_go_bar2 <- ggplot(go_df2, aes(x = log10_fe, y = term_ordered, fill = p_value)) +
    geom_col(width = 0.75, colour = "grey30", linewidth = 0.2) +
    scale_fill_gradient(
      low  = "#08306B", high = "#9ECAE1",
      name = "p-value"
    ) +
    ## Strip the "source___" prefix from y-axis labels
    scale_y_discrete(labels = function(x) gsub("^.*___", "", x)) +
    facet_grid(source_label ~ ., scales = "free_y", space = "free_y") +
    labs(
      title = "Selected GO Terms",
      x     = expression(log[10](Fold~Enrichment)),
      y     = NULL
    ) +
    theme_publication() +
    theme(
      strip.text.y.right  = ggtext::element_textbox(size=9,face= "bold",orientation = "upright",width= unit(2, "cm"),halign= 0.5,valign= 0.5,margin= margin(1, 1, 1,1)),
      axis.text.y        = element_text(size = 10, hjust = 1),
      plot.title = element_text(margin = margin(b = 8))
    )

  print(p_go_bar2)
}
```


## Pathway Enrichment — Bar Charts {.tabset}

### KEGG Pathways

```{r kegg-bar, fig.height=3, fig.width=10}
################################################################################
# KEGG pathway bar chart — same style as the GO bar chart.
################################################################################

kegg_df <- plot_df %>% dplyr::filter(source == "KEGG")

if (nrow(kegg_df) > 0) {
  p_kegg <- ggplot(
    kegg_df %>% dplyr::mutate(term_wrap = fct_reorder(term_wrap, log10_fe)),
    aes(x = log10_fe, y = term_wrap, fill = p_value)
  ) +
    geom_col(width = 0.7, colour = "grey30", linewidth = 0.2) +
    scale_fill_gradient(low = "#006D2C", high = "#A1D99B", name = "p-value") +
    labs(
      title = "KEGG Pathway Enrichment",
      x     = expression(log[10](Fold~Enrichment)),
      y     = NULL
    ) +
    theme_publication() +
    theme(axis.text.y = element_text(hjust = 1, margin = margin(2,2,2,2)),
          plot.title = element_text(margin = margin(b = 8)))

  print(p_kegg)
} else {
  cat("No KEGG terms found.\n")
}
```

### Reactome Pathways

```{r reactome-bar, fig.height=6, fig.width=10}
################################################################################
# Reactome pathway bar chart.
################################################################################

reac_df <- plot_df %>% dplyr::filter(source == "REAC")

if (nrow(reac_df) > 0) {
  p_reac <- ggplot(
    reac_df %>% dplyr::mutate(term_wrap = fct_reorder(term_wrap, log10_fe)),
    aes(x = log10_fe, y = term_wrap, fill = p_value)
  ) +
    geom_col(width = 0.7, colour = "grey30", linewidth = 0.2) +
    scale_fill_gradient(low = "#7A0177", high = "#FEEBE2", name = "p-value") +
    labs(
      title = "Reactome Pathway Enrichment",
      x     = expression(log[10](Fold~Enrichment)),
      y     = NULL
    ) +
    theme_publication() +
    theme(axis.text.y = element_text(hjust = 1, margin = margin(2,2,2,2)),
          plot.title = element_text(margin = margin(b = 8)))

  print(p_reac)
} else {
  cat("No Reactome terms found.\n")
}
```

### WikiPathways

```{r wp-bar, fig.height=6, fig.width=10}
################################################################################
# WikiPathways bar chart.
################################################################################

wp_df <- plot_df %>% dplyr::filter(source == "WP")

if (nrow(wp_df) > 0) {
  p_wp <- ggplot(
    wp_df %>% dplyr::mutate(term_wrap = fct_reorder(term_wrap, log10_fe)),
    aes(x = log10_fe, y = term_wrap, fill = p_value)
  ) +
    geom_col(width = 0.7, colour = "grey30", linewidth = 0.2) +
    scale_fill_gradient(low = "#B35806", high = "#FEE0B6", name = "p-value") +
    labs(
      title = "WikiPathways Enrichment",
      x     = expression(log[10](Fold~Enrichment)),
      y     = NULL
    ) +
    theme_publication() +
    theme(axis.text.y = element_text(hjust = 1, margin = margin(2,2,2,2)),
          plot.title = element_text(margin = margin(b = 8)))

  print(p_wp)
} else {
  cat("No WikiPathways terms found.\n")
}
```


## All Sources Combined — Dot Plot

Dot plots (bubble plots) encode three variables simultaneously:
x-axis = gene ratio, dot size = intersection count, colour = p-value.

```{r dotplot-all, fig.height=10, fig.width=12}
################################################################################
# Combined dot plot across all enrichment sources.
# - x-axis : gene_ratio (intersection_size / query_size)
# - y-axis : term name (ordered by gene_ratio within each source)
# - size   : intersection_size (how many query genes hit this term)
# - colour : p_value (darker = more significant)
################################################################################

dot_df <- plot_df %>%
  dplyr::arrange(source_label, gene_ratio) %>%
  dplyr::mutate(
    term_ordered = factor(paste(source, term_wrap, sep = "___"),
                          levels = paste(source, term_wrap, sep = "___"))
  )

if (nrow(dot_df) > 0) {
  p_dot <- ggplot(dot_df, aes(x = gene_ratio, y = term_ordered,
                               size = intersection_size, colour = p_value)) +
    geom_segment(aes(x = 0, xend = gene_ratio, y = term_ordered, yend = term_ordered),
             linewidth = 0.5, colour = "grey60") +
    geom_point(alpha = 1) +
    scale_colour_gradient(
      low  = "#08306B", high = "#9ECAE1",
      name = "p-value"
    ) +
    scale_size_continuous(
      range = c(2, 8),
      name  = "Gene count"
    ) +
    scale_y_discrete(labels = function(x) gsub("^.*___", "", x)) +
    facet_grid(source_label ~ ., scales = "free_y", space = "free_y") +
    labs(
      title = "Enrichment Dot Plot — All Sources",
      x     = "Gene Ratio (intersection / query)",
      y     = NULL
    ) +
    theme_publication() +
    theme(
      strip.text.y.right  = ggtext::element_textbox(size=9,face= "bold",orientation = "upright",width= unit(2, "cm"),halign= 0.5,valign= 0.5,margin= margin(1, 1, 1,1)),
      axis.text.y        = element_text(hjust = 1),
      plot.title = element_text(margin = margin(b = 8))
    )

  print(p_dot)
}
```


## Lollipop Plot — Top Enriched Terms

Lollipop plots provide a cleaner alternative to bar charts,
especially when many terms are shown.

```{r lollipop, fig.height=9, fig.width=11}
################################################################################
# Lollipop chart: each term is a point connected to the y-axis by a segment.
# Colour = p-value, x-axis = -log10(p_value).
################################################################################

lollipop_df <- plot_df %>%
  dplyr::arrange(source_label, neg_log10_p) %>%
  dplyr::mutate(
    term_ordered = factor(paste(source, term_wrap, sep = "___"),
                          levels = paste(source, term_wrap, sep = "___"))
  )

if (nrow(lollipop_df) > 0) {
  p_lollipop <- ggplot(lollipop_df, aes(x = neg_log10_p, y = term_ordered,
                                         colour = p_value)) +
    ## Stem
    geom_segment(aes(xend = 0, yend = term_ordered),
                 linewidth = 0.6, colour = "grey60") +
    ## Point
    geom_point(aes(size = intersection_size), alpha = 1) +
    scale_colour_gradient(low = "#08306B", high = "#9ECAE1", name = "p-value") +
    scale_size_continuous(range = c(2.5, 7), name = "Gene count") +
    scale_y_discrete(labels = function(x) gsub("^.*___", "", x)) +
    facet_grid(source_label ~ ., scales = "free_y", space = "free_y") +
    labs(
      title = "Enrichment Lollipop Plot — All Sources",
      x     = expression(-log[10](italic(p))),
      y     = NULL
    ) +
    theme_publication() +
    theme(
      strip.text.y.right  = ggtext::element_textbox(size=9,face= "bold",orientation = "upright",width= unit(2, "cm"),halign= 0.5,valign= 0.5,margin= margin(1, 1, 1,1)),
      axis.text.y        = element_text(hjust = 1),
      plot.title = element_text(margin = margin(b = 8))
    )

  print(p_lollipop)
}
```


## Bar Chart — Fold Enrichment (All Sources)

A combined faceted bar chart for all available sources, matching the
reference figure styling.

```{r bar-all-sources, fig.height=12, fig.width=12}
################################################################################
# Bar chart of top terms across ALL sources, faceted with uniform bar thickness.
# Colour fill = p-value, x-axis = log10(Fold Enrichment).
#
# Key layout choices:
#   facet_grid(... , space = "free_y")  — each panel's height is proportional
#     to its number of terms, so bars keep the same thickness everywhere.
#     (facet_wrap cannot do this — it gives equal panel heights, stretching
#     bars in panels with fewer terms.)
#   hjust = 1  — right-aligns y-axis labels so they sit close to the axis
#     rather than floating left-justified.
################################################################################

bar_all_df <- plot_df %>%
  dplyr::arrange(source_label, log10_fe) %>%
  dplyr::mutate(
    term_ordered = factor(paste(source, term_wrap, sep = "___"),
                          levels = paste(source, term_wrap, sep = "___"))
  )

if (nrow(bar_all_df) > 0) {
  p_bar_all <- ggplot(bar_all_df, aes(x = log10_fe, y = term_ordered,
                                       fill = p_value)) +
    geom_col(width = 0.7, colour = "grey30", linewidth = 0.2) +
    scale_fill_gradient(
      low  = "#08306B", high = "#9ECAE1",
      name = "p-value",
      guide = guide_colorbar(reverse = FALSE)
    ) +
    scale_y_discrete(labels = function(x) gsub("^.*___", "", x)) +
    ## facet_grid with space = "free_y" keeps bar thickness uniform
    facet_grid(source_label ~ ., scales = "free_y", space = "free_y") +
    labs(
      title = "Enrichment Bar Chart — All Sources",
      x     = expression(log[10](Fold~Enrichment)),
      y     = NULL
    ) +
    theme_publication() +
    theme(
      strip.text.y.right  = ggtext::element_textbox(size=9,face= "bold",orientation = "upright",width= unit(2, "cm"),halign= 0.5,valign= 0.5,margin= margin(1, 1, 1,1)),
      axis.text.y         = element_text(size = 9, hjust = 1),
      plot.title = element_text(margin = margin(b = 8))
    )

  print(p_bar_all)
}
```


## Heatmap-Style — Gene Ratio by Source

```{r heatmap-tile, fig.height=8, fig.width=12}
################################################################################
# Tile / heatmap plot showing gene ratio across top terms.
# Each tile is coloured by -log10(p-value); useful for comparing enrichment
# strength across ontology categories.
################################################################################

## Take top 5 per source for a compact heatmap
heat_df <- enrich_df %>%
  dplyr::group_by(source) %>%
  dplyr::slice_min(p_value, n = 5, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    source_label = ifelse(source %in% names(source_labels),
                          source_labels[source], source),
    term_short   = stringr::str_trunc(term_name, 50, side = "right")
  ) %>%
  ## Order terms: group by source, then by significance within each source.
  ## This prevents terms from different sources being interleaved on the y-axis.
  dplyr::arrange(source, neg_log10_p) %>%
  dplyr::mutate(
    term_short = factor(term_short, levels = unique(term_short))
  )

if (nrow(heat_df) > 0) {
  p_heat <- ggplot(heat_df, aes(x = source_label,
                                 y = term_short,
                                 fill = neg_log10_p)) +
    geom_tile(colour = "white", linewidth = 0.5) +
    geom_text(aes(label = intersection_size), colour = "white",
              fontface = "bold", size = 3.5) +
    scale_fill_distiller(palette = "YlOrRd", direction = 1, name = expression(-log[10](p))) +
    labs(
      title = "Enrichment Heatmap — Top 5 Terms per Source",
      x     = NULL,
      y     = NULL
    ) +
    theme_publication() +
    theme(
      axis.text.x     = element_text(angle = 30, hjust = 1, size = 10),
      axis.text.y     = element_text(size = 9, hjust = 1, margin = margin(2,2,2,2)),
      panel.grid       = element_blank(),
      panel.background = element_rect(fill = "grey95"),
      plot.title = element_text(margin = margin(b = 8))
    )

  print(p_heat)
}
```


## Intersection Size Bar Chart

```{r intersection-bar, fig.height=7, fig.width=10}
################################################################################
# Simple bar chart showing the number of overlapping genes per term,
# coloured by source. Useful for identifying terms with the largest gene hits.
################################################################################

count_df <- plot_df %>%
  dplyr::arrange(desc(intersection_size)) %>%
  dplyr::slice_head(n = 25) %>%
  dplyr::mutate(term_short = stringr::str_trunc(term_name, 50, side = "right"))

if (nrow(count_df) > 0) {
  ## Source colour palette
  src_colours <- c(
    "GO:BP" = "#1B9E77", "GO:CC" = "#D95F02", "GO:MF" = "#7570B3",
    "KEGG"  = "#E7298A", "REAC"  = "#66A61E", "WP"    = "#E6AB02",
    "TF"    = "#A6761D", "MIRNA" = "#666666", "CORUM" = "#1F78B4",
    "HPA"   = "#FB9A99", "HP"    = "#CAB2D6"
  )

  p_count <- ggplot(count_df,
                    aes(x = intersection_size,
                        y = reorder(term_short, intersection_size),
                        fill = source)) +
    geom_col(width = 0.7) +
    scale_fill_manual(values = src_colours, name = "Source") +
    labs(
      title = "Top 25 Enriched Terms by Gene Count",
      x     = "Number of Overlapping Genes",
      y     = NULL
    ) +
    theme_publication() + 
    theme(axis.text.y = element_text(hjust = 1, margin = margin(2,2,2,2)),
          plot.title = element_text(margin = margin(b = 8)))

  print(p_count)
}
```


# Save Publication Figures

```{r save-figures, eval=FALSE}
################################################################################
# Save key figures as high-resolution PNG and PDF for publication.
# Set eval=TRUE to run this chunk, or run manually.
# Adjust width, height, and dpi as needed for journal requirements.
################################################################################

## GO combined bar chart (tidytext version)
if (exists("p_go_bar")) {
  ggsave("GO_terms_bar_chart.png",  p_go_bar, width = 10, height = 9,  dpi = 300)
  ggsave("GO_terms_bar_chart.pdf",  p_go_bar, width = 10, height = 9)
}

## GO combined bar chart (alternative, no tidytext)
if (exists("p_go_bar2")) {
  ggsave("GO_terms_bar_chart_alt.png",  p_go_bar2, width = 10, height = 9,  dpi = 300)
  ggsave("GO_terms_bar_chart_alt.pdf",  p_go_bar2, width = 10, height = 9)
}

## Dot plot
if (exists("p_dot")) {
  ggsave("enrichment_dotplot.png",  p_dot, width = 11, height = 10, dpi = 300)
  ggsave("enrichment_dotplot.pdf",  p_dot, width = 11, height = 10)
}

## Lollipop plot
if (exists("p_lollipop")) {
  ggsave("enrichment_lollipop.png", p_lollipop, width = 10, height = 9, dpi = 300)
  ggsave("enrichment_lollipop.pdf", p_lollipop, width = 10, height = 9)
}

## Bar chart — all sources
if (exists("p_bar_all")) {
  ggsave("enrichment_bar_all_sources.png", p_bar_all, width = 11, height = 12, dpi = 300)
  ggsave("enrichment_bar_all_sources.pdf", p_bar_all, width = 11, height = 12)
}

## KEGG
if (exists("p_kegg")) {
  ggsave("KEGG_pathway_bar.png", p_kegg, width = 10, height = 6, dpi = 300)
  ggsave("KEGG_pathway_bar.pdf", p_kegg, width = 10, height = 6)
}

## Reactome
if (exists("p_reac")) {
  ggsave("Reactome_pathway_bar.png", p_reac, width = 10, height = 6, dpi = 300)
  ggsave("Reactome_pathway_bar.pdf", p_reac, width = 10, height = 6)
}

## WikiPathways
if (exists("p_wp")) {
  ggsave("WikiPathways_bar.png", p_wp, width = 10, height = 6, dpi = 300)
  ggsave("WikiPathways_bar.pdf", p_wp, width = 10, height = 6)
}

cat("All figures saved.\n")
```


# Session Information

```{r session-info}
################################################################################
# Record the session info for reproducibility.
# Includes R version, platform, loaded packages and their versions.
################################################################################
sessionInfo()
```

---

**End of analysis.** All results are exported as TSV/CSV files in the
working directory. The HTML report contains interactive tables and both
interactive (plotly) and static (ggplot2) figures.
