---
title: "GO & Pathway Enrichment Analysis for Non-Model Organisms"
subtitle: "*Pangasianodon hypophthalmus* (Striped Catfish) — GO (enricher, topGO), KEGG, Reactome, WikiPathways & PANTHER"
author: "Phuong Nguyen"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 2
    number_sections: true
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 10,
  fig.height = 7,
  fig.align  = "center",
  cache      = FALSE          # no caching — every run is reproducible from scratch
)
```

# Overview

This pipeline performs Gene Ontology (GO) and pathway over-representation
analysis (ORA) for a non-model organism (*Pangasianodon hypophthalmus*, striped
catfish, NCBI TaxID **310915**) from a differentially-expressed (DE) gene list.

The workflow is organised into six stages:

1. **Setup** — all user-configurable parameters, packages, and helper functions.
2. **Input data** — load the pre-built annotation files and the DE / background
   gene lists.
3. **GO enrichment** — `clusterProfiler::enricher()` for BP / MF / CC, plus
   DAG-aware `topGO`.
4. **Pathway enrichment** — KEGG (direct) and Reactome / WikiPathways / PANTHER
   (via zebrafish orthologs).
5. **Validation & benchmarking** — a manual hypergeometric test that reproduces
   `enricher()`, and cross-method / cross-database comparisons.
6. **Summary** — the parameters used and a single at-a-glance count of every
   enrichment result.

> This document assumes the annotation files already exist in the working
> directory. See **Input data → Obtaining the annotation files** for one-time
> `bash` preparation instructions.

---

# Setup

## Analysis parameters {#parameters}

**Everything a user needs to change lives in this one chunk.** Edit the values
here; nothing below needs to be touched.

```{r params}
## ── Organism ────────────────────────────────────────────────────────────────
TAXID          <- 310915
SPECIES_NAME   <- "Pangasianodon hypophthalmus"

## ── Input files (must already exist in the working directory) ────────────────
GENE2GO_FILE    <- "Phypophthalmus_gene2go.tsv"    # organism GO annotations (required)
GENE_INFO_FILE  <- "Phypophthalmus_gene_info.tsv"  # symbol ↔ ID lookup (needed only if DE list uses symbols)
DE_GENE_FILE    <- "de_gene_list_symbols.txt"      # DE genes: NCBI IDs OR symbols, one per line
BACKGROUND_FILE <- "genelist"                      # custom background file path, or NULL = all annotated genes
ORTHO_FILE      <- "Phypophthalmus_to_drerio_orthologs.tsv"  # cached ortholog map (auto-created if absent)

## ── Statistical thresholds ───────────────────────────────────────────────────
PVALUE_CUTOFF <- 0.05     # adjusted-p cutoff for reporting / calling significance
QVALUE_CUTOFF <- 0.2      # q-value cutoff (enricher/enrichKEGG/etc.)
PADJ_METHOD   <- "BH"     # multiple-testing correction ("BH", "holm", "none", ...)

## ── Gene-set eligibility filters ─────────────────────────────────────────────
MIN_DE_IN_TERM <- 2       # ≥ this many DE genes must map to a term  (Count filter)
MIN_GENE_SET   <- 5       # ≥ this many background genes per term     (minGSSize / topGO nodeSize)
MAX_GENE_SET   <- 500     # ≤ this many background genes per term     (maxGSSize)

## ── Which analyses to run ────────────────────────────────────────────────────
RUN_TOPGO              <- TRUE   # DAG-aware GO enrichment (requires topGO)
RUN_PATHWAY_ENRICHMENT <- TRUE   # KEGG + ortholog-based pathway databases

## ── KEGG ─────────────────────────────────────────────────────────────────────
# If YOUR species is a registered KEGG organism, set KEGG_ORG to its code and
# keep KEGG_VIA_ORTHOLOGS = FALSE (runs directly on your NCBI IDs). If it is NOT
# in KEGG, set KEGG_ORG to the REFERENCE organism's code (e.g. "gga" chicken,
# "dre" zebrafish, "hsa" human) and KEGG_VIA_ORTHOLOGS = TRUE (runs on orthologs).
KEGG_ORG           <- "phyp"     # KEGG organism code (see https://www.genome.jp/kegg/catalog/org_list.html)
KEGG_VIA_ORTHOLOGS <- FALSE      # TRUE = run KEGG on the reference-organism orthologs

## ── Pathway databases that need a model-organism reference ───────────────────
# Reactome / WikiPathways / PANTHER cover only model organisms, so your genes are
# mapped to a REFERENCE organism (see the "ortholog mapping" parameters below).
# The value here MUST be an organism each database supports:
#   • REACTOME_ORG : one of ReactomePA's set — "human", "mouse", "rat",
#       "zebrafish", "fly", "celegans", "yeast". (Chicken is NOT supported; for
#       birds map to "human" for Reactome.) Check: ReactomePA supported organisms.
#   • WP_ORGANISM  : a WikiPathways species; check clusterProfiler::get_wp_organisms()
#   • PANTHER_TAXID: a PANTHER reference TaxID; check rbioapi::rba_panther_info()
REACTOME_ORG   <- "zebrafish"    # Reactome organism (ortholog target)
WP_ORGANISM    <- "Danio rerio"  # WikiPathways organism (ortholog target)
PANTHER_TAXID  <- 7955           # PANTHER organism NCBI TaxID (D. rerio = 7955; chicken = 9031)

## ── Ortholog mapping: YOUR species → the REFERENCE organism ──────────────────
# These drive the ortholog-mapping chunk. Change them to retarget the reference
# organism (e.g. a bird → chicken). See "Adapting to another species" below.
#   ORTHO_ORGDB           : Bioconductor annotation package for the REFERENCE
#                           organism (zebrafish "org.Dr.eg.db", chicken
#                           "org.Gg.eg.db", human "org.Hs.eg.db", mouse "org.Mm.eg.db").
#   ORTHO_BIOMART_PREFIX  : Ensembl homolog-attribute prefix for the reference
#                           (zebrafish "drerio", chicken "ggallus", human "hsapiens").
#   ORTHO_ENSEMBL_DATASET : Ensembl gene dataset for the reference
#                           ("drerio_gene_ensembl", "ggallus_gene_ensembl", ...).
#   SPECIES_BIOMART_REGEX : regex that matches YOUR species' Ensembl dataset name
#                           (leave "" to skip BioMart and use symbol matching only).
ORTHO_ORGDB           <- "org.Dr.eg.db"
ORTHO_BIOMART_PREFIX  <- "drerio"
ORTHO_ENSEMBL_DATASET <- "drerio_gene_ensembl"
SPECIES_BIOMART_REGEX <- "hypophthalmus|pangasianodon"

## ── Display / output ─────────────────────────────────────────────────────────
SHOW_CATEGORY <- 20              # number of terms shown in per-analysis plots
OUT_DIR       <- "GO_results"    # all TSVs and PNGs are written here

ontology_colours <- c("BP" = "#E64B35", "MF" = "#4DBBD5", "CC" = "#00A087")
pathway_colours  <- c("KEGG" = "#E64B35", "Reactome" = "#4DBBD5",
                      "WikiPathways" = "#00A087", "PANTHER" = "#F39B7F")

## ── Sanity checks on the parameters above ───────────────────────────────────
stopifnot(
  MIN_DE_IN_TERM >= 1,
  MIN_GENE_SET   >= 1,
  MAX_GENE_SET   >= MIN_GENE_SET,
  PVALUE_CUTOFF > 0, PVALUE_CUTOFF <= 1,
  QVALUE_CUTOFF > 0, QVALUE_CUTOFF <= 1
)

## ── Output directories ──────────────────────────────────────────────────────
PLOT_DIR         <- file.path(OUT_DIR, "plots")
PATHWAY_DIR      <- file.path(OUT_DIR, "pathway_results")
PATHWAY_PLOT_DIR <- file.path(PLOT_DIR, "pathways")
for (d in c(OUT_DIR, PLOT_DIR, PATHWAY_DIR, PATHWAY_PLOT_DIR))
  dir.create(d, showWarnings = FALSE, recursive = TRUE)
```

## Packages

```{r install-packages, eval=FALSE}
# ── Run once ──
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install(c("clusterProfiler", "enrichplot", "GO.db", "AnnotationDbi",
                       "topGO", "ReactomePA", "DOSE",
                       "org.Dr.eg.db",   # reference-organism annotation pkg — swap to match
                                         # ORTHO_ORGDB, e.g. org.Gg.eg.db (chicken), org.Hs.eg.db (human)
                       "biomaRt", "KEGGREST"))
install.packages(c("tidyverse", "DT", "scales", "ggrepel", "rbioapi"))
```

```{r load-packages}
library(tidyverse)
library(clusterProfiler)
library(enrichplot)
library(GO.db)
library(AnnotationDbi)
library(DT)
library(scales)
library(ggrepel)

# Resolve namespace conflicts (AnnotationDbi::select / IRanges::slice mask dplyr)
select <- dplyr::select
slice  <- dplyr::slice
```

## Helper functions

Small utilities reused throughout the pipeline: the gene-set overlap filter, a
ratio parser, generic table/plot renderers, and the symbol→ID converter.

```{r helpers}
# ── Enforce the ≥ MIN_DE_IN_TERM overlap filter on an enrichResult ───────────
# enricher()/enrichKEGG()/enrichPathway()/enrichWP() can report a term hit by a
# single DE gene; they expose no minimum-overlap argument. We drop terms whose
# DE-gene overlap (Count) is below the threshold. Returns the object unchanged
# if it is NULL or lacks a Count column.
filter_min_overlap <- function(res, min_k = MIN_DE_IN_TERM) {
  if (is.null(res)) return(res)
  if (!"Count" %in% colnames(res@result)) return(res)
  res@result <- res@result[!is.na(res@result$Count) &
                             res@result$Count >= min_k, , drop = FALSE]
  res
}

# ── Parse a "k/n" GeneRatio/BgRatio string into a numeric proportion ─────────
parse_ratio <- function(ratio_str) {
  parts <- str_split(ratio_str, "/", simplify = TRUE)
  as.numeric(parts[, 1]) / as.numeric(parts[, 2])
}

# ── TRUE if an enrichResult has ≥ n rows ─────────────────────────────────────
has_rows <- function(res, n = 1) !is.null(res) && nrow(as.data.frame(res)) >= n

# ── Number of significant terms (padj < cutoff) in an enrichResult ───────────
n_sig <- function(res, cutoff = PVALUE_CUTOFF) {
  if (!has_rows(res)) return(0L)
  sum(as.data.frame(res)$p.adjust < cutoff, na.rm = TRUE)
}

# ── Standard results table for an enrichResult ───────────────────────────────
enrich_table <- function(res, caption) {
  if (!has_rows(res)) { cat("No results to display.\n"); return(invisible()) }
  as.data.frame(res) %>%
    dplyr::select(ID, Description, GeneRatio, BgRatio, pvalue, p.adjust, qvalue, Count) %>%
    mutate(across(c(pvalue, p.adjust, qvalue), ~ signif(.x, 3))) %>%
    datatable(rownames = FALSE, caption = caption,
              options = list(pageLength = 15, scrollX = TRUE))
}

# ── Dot plot for an enrichResult ─────────────────────────────────────────────
enrich_dotplot <- function(res, title) {
  dotplot(res, showCategory = SHOW_CATEGORY, font.size = 11) +
    ggtitle(title) + theme(plot.title = element_text(face = "bold", size = 14))
}

# ── Custom horizontal bar plot (enrichplot::barplot was removed upstream) ────
enrich_barplot <- function(res, showCategory = SHOW_CATEGORY, title = NULL) {
  df <- as.data.frame(res) %>%
    head(showCategory) %>%
    mutate(Description = factor(Description, levels = rev(Description)))
  ggplot(df, aes(x = Count, y = Description, fill = p.adjust)) +
    geom_col(width = 0.7) +
    scale_fill_gradient(low = "#E64B35", high = "#4292C6",
                        name = "p.adjust", trans = "log10") +
    labs(x = "Gene count", y = NULL, title = title) +
    theme_minimal(base_size = 12) +
    theme(plot.title = element_text(face = "bold", size = 14))
}

# ── Save a plot safely (build inside tryCatch so failures never halt knit) ───
safe_save <- function(filename, plot_fn, dir = PLOT_DIR, w = 10, h = 8) {
  tryCatch({
    ggsave(file.path(dir, filename), plot = plot_fn(),
           width = w, height = h, dpi = 300, bg = "white")
    message("Saved: ", filename)
  }, error = function(e) message("Skipped ", filename, ": ", e$message))
}

# ── Convert gene symbols → NCBI Gene IDs using the gene_info lookup ───────────
convert_symbols_to_ids <- function(symbols, gene_info) {
  lookup <- gene_info %>% mutate(Symbol_lower = tolower(Symbol))
  symbols_lower <- tolower(trimws(symbols))
  match_df <- tibble(input = symbols, input_lower = symbols_lower) %>%
    left_join(lookup %>% dplyr::select(GeneID, Symbol, Symbol_lower),
              by = c("input_lower" = "Symbol_lower"))

  matched   <- match_df %>% filter(!is.na(GeneID))
  unmatched <- match_df %>% filter(is.na(GeneID))

  # Fall back to the Synonyms column for anything still unmatched
  if (nrow(unmatched) > 0 && "Synonyms" %in% colnames(gene_info)) {
    synonym_lookup <- gene_info %>%
      filter(Synonyms != "-") %>%
      mutate(syn_list = strsplit(Synonyms, "\\|")) %>%
      unnest(syn_list) %>%
      mutate(syn_lower = tolower(trimws(syn_list))) %>%
      dplyr::select(GeneID, Symbol, syn_lower)

    synonym_match <- unmatched %>%
      dplyr::select(input, input_lower) %>%
      left_join(synonym_lookup, by = c("input_lower" = "syn_lower"))

    matched <- bind_rows(
      matched %>% dplyr::select(input, GeneID, Symbol),
      synonym_match %>% filter(!is.na(GeneID)) %>% dplyr::select(input, GeneID, Symbol)
    )
    unmatched <- synonym_match %>% filter(is.na(GeneID))
  }

  cat(sprintf("Symbol conversion: %d / %d symbols matched to NCBI Gene IDs\n",
              n_distinct(matched$input), length(symbols)))
  if (nrow(unmatched) > 0) {
    cat(sprintf("  %d not found: %s%s\n", nrow(unmatched),
                paste(head(unmatched$input, 20), collapse = ", "),
                if (nrow(unmatched) > 20) " ..." else ""))
  }
  # One symbol → several IDs: keep the first
  matched <- matched %>% group_by(input) %>% dplyr::slice(1) %>% ungroup()

  list(ids = matched$GeneID, matched = matched, unmatched = unmatched$input)
}
```

---

# Input data

## Obtaining the annotation files (one-time, `bash`)

This script **loads** pre-built, organism-filtered files; it does not download
anything. Prepare the three files below once with the commands shown, then
point the parameters in [Analysis parameters](#parameters) at them.

**1. GO annotations — `Phypophthalmus_gene2go.tsv` (required).**
Filter NCBI's genome-wide `gene2go` to your TaxID. Keep the header.

```bash
wget https://ftp.ncbi.nlm.nih.gov/gene/DATA/gene2go.gz
gunzip gene2go.gz
# Keep the header line (NR==1) plus rows for TaxID 310915 (column 1)
awk -F'\t' 'NR==1 || $1==310915' gene2go > Phypophthalmus_gene2go.tsv
# Expected columns: TaxID GeneID GO_ID Evidence Qualifier GO_term PubMed Category
```

**2. Symbol ↔ ID lookup — `Phypophthalmus_gene_info.tsv` (only if your DE list uses symbols).**

```bash
wget https://ftp.ncbi.nlm.nih.gov/gene/DATA/gene_info.gz
gunzip gene_info.gz
awk -F'\t' 'NR==1 || $1==310915' gene_info > gene_info_sp.txt
# Then keep just the useful columns (GeneID, Symbol, Synonyms, description, type_of_gene)
# with cut/csvcut, saving as Phypophthalmus_gene_info.tsv
```

**3. Orthologs to the reference organism — `ORTHO_FILE` (only for Reactome / WikiPathways / PANTHER, or KEGG-via-orthologs).**
If this file is absent, the pipeline builds it automatically (Ensembl BioMart,
falling back to symbol matching against `ORTHO_ORGDB`) and caches it. You can
also build it yourself; the pipeline only needs two columns, **`sp_geneid`**
(your species' NCBI GeneID) and **`ref_entrez`** (the reference organism's NCBI
GeneID). The fastest offline route is NCBI's `gene_orthologs`:

```bash
wget https://ftp.ncbi.nlm.nih.gov/gene/DATA/gene_orthologs.gz
gunzip gene_orthologs.gz
# Columns: tax_id  GeneID  relationship  Other_tax_id  Other_GeneID
# Keep pairs between YOUR species (310915) and the reference (zebrafish = 7955),
# in EITHER direction, and emit a 2-column sp_geneid <TAB> ref_entrez table:
{ echo -e "sp_geneid\tref_entrez"
  awk -F'\t' '$1==310915 && $4==7955 {print $2"\t"$5}' gene_orthologs
  awk -F'\t' '$1==7955 && $4==310915 {print $5"\t"$2}' gene_orthologs
} | sort -u > Phypophthalmus_to_drerio_orthologs.tsv
```

> After building the small filtered files you can delete the large genome-wide
> downloads — only the per-organism TSVs are used here.

## Adapting to another species {#adapting}

The pipeline is not catfish-specific. To run it on **any** species — say a
non-model **bird** whose pathways you want against **chicken (*Gallus gallus*)** —
change only the parameters, not the code. Two things drive everything: your
species' TaxID, and the **reference organism** each pathway database is mapped to.

**Step 1 — Rebuild the GO / symbol files for your species** exactly as in items
1–2 above, substituting your TaxID for `310915`.

**Step 2 — Pick a reference organism supported by each database.** This is the
crucial step: the reference you map to *must* be one the database actually
covers. Coverage differs by database, so a single reference may not serve all
four:

| Database | How to check supported organisms | Chicken? |
|---|---|---|
| KEGG | <https://www.genome.jp/kegg/catalog/org_list.html> (code e.g. `gga`) | ✅ `gga` |
| Reactome (`ReactomePA`) | fixed set: human, mouse, rat, **zebrafish**, fly, celegans, yeast | ❌ — use **human** |
| WikiPathways | `clusterProfiler::get_wp_organisms()` | ✅ `Gallus gallus` |
| PANTHER | `rbioapi::rba_panther_info()` organism list | ✅ TaxID `9031` |

Because chicken is **not** in `ReactomePA`, a bird analysis typically maps to
**chicken** for KEGG / WikiPathways / PANTHER, but to **human** (or zebrafish)
for Reactome. You can either set `REACTOME_ORG <- "human"` and build a
*second* ortholog file to human, or simply skip Reactome. If one reference
covers all your target databases, one ortholog file is enough.

**Step 3 — Set the parameters** (top of the document). For a bird → chicken run:

```r
TAXID                 <- <your_bird_taxid>
SPECIES_NAME          <- "<Your bird species>"
GENE2GO_FILE          <- "<yourbird>_gene2go.tsv"
GENE_INFO_FILE        <- "<yourbird>_gene_info.tsv"
ORTHO_FILE            <- "<yourbird>_to_ggallus_orthologs.tsv"

# KEGG: if your bird IS in KEGG, use its own code and keep the toggle FALSE.
# If it is NOT, map to chicken:
KEGG_ORG              <- "gga"        # chicken KEGG code
KEGG_VIA_ORTHOLOGS    <- TRUE         # run KEGG on chicken orthologs

# Reactome cannot use chicken → map Reactome to human (or set to skip)
REACTOME_ORG          <- "human"
WP_ORGANISM           <- "Gallus gallus"
PANTHER_TAXID         <- 9031          # chicken

# Ortholog target = chicken (used for WikiPathways / PANTHER / KEGG-orthologs)
ORTHO_ORGDB           <- "org.Gg.eg.db"       # BiocManager::install("org.Gg.eg.db")
ORTHO_BIOMART_PREFIX  <- "ggallus"            # Ensembl homolog attribute prefix
ORTHO_ENSEMBL_DATASET <- "ggallus_gene_ensembl"
SPECIES_BIOMART_REGEX <- "<yourbird_ensembl_pattern>"  # e.g. "taeniopygia|zebra finch"; "" to skip BioMart
```

> **Mismatched references caveat.** If Reactome uses human orthologs while
> WikiPathways/PANTHER use chicken, their backgrounds differ and cross-database
> comparisons in the Summary are only qualitative. That is expected — enrichment
> is always relative to its own universe.

**Step 4 — Build the ortholog file.** Easiest is to let the pipeline build it
(needs `biomaRt` + your `ORTHO_ORGDB`); it auto-caches to `ORTHO_FILE`. To build
it offline for a bird → chicken (TaxID `9031`) map, reuse the `gene_orthologs`
recipe above with your TaxID and `9031`, writing the `sp_geneid`/`ref_entrez`
columns. Finding your species' Ensembl dataset name for `SPECIES_BIOMART_REGEX`:

```r
library(biomaRt); ens <- useMart("ensembl")
# search the dataset list for your species keyword:
subset(listDatasets(ens), grepl("finch|taeniopygia", dataset, ignore.case = TRUE))
```

If your species is **not in Ensembl**, set `SPECIES_BIOMART_REGEX <- ""` to skip
BioMart; the pipeline then matches your `gene_info` symbols against the reference
`ORTHO_ORGDB` symbols (lower-quality but dependency-free). For the best coverage
on truly non-model species, generate orthologs with **eggNOG-mapper** or
**OrthoFinder** and save them as the two-column `sp_geneid`/`ref_entrez` TSV.

**Step 5 — Run.** No chunk edits are required: the ortholog-mapping and pathway
chunks read `ORTHO_ORGDB`, `ORTHO_BIOMART_PREFIX`, `ORTHO_ENSEMBL_DATASET`,
`REACTOME_ORG`, `WP_ORGANISM`, `PANTHER_TAXID`, `KEGG_ORG` and
`KEGG_VIA_ORTHOLOGS` directly from the parameters above.

## Load GO annotations

```{r load-gene2go}
if (!file.exists(GENE2GO_FILE))
  stop("GO annotation file not found: ", GENE2GO_FILE,
       "\n  See 'Obtaining the annotation files' above.")

gene2go_sp <- read.delim(GENE2GO_FILE, stringsAsFactors = FALSE)

# Guard against a headerless / mis-columned file
required_cols <- c("GeneID", "GO_ID", "GO_term", "Category")
if (!all(required_cols %in% colnames(gene2go_sp)))
  stop("`", GENE2GO_FILE, "` is missing expected columns: ",
       paste(setdiff(required_cols, colnames(gene2go_sp)), collapse = ", "))

cat(sprintf(
  "GO annotations for %s (TaxID %d):\n  %s genes | %s GO terms | %s associations\n",
  SPECIES_NAME, TAXID,
  comma(n_distinct(gene2go_sp$GeneID)),
  comma(n_distinct(gene2go_sp$GO_ID)),
  comma(nrow(gene2go_sp))
))
```

```{r annotation-eda, fig.height=4.5}
# Coverage by ontology
gene2go_sp %>%
  group_by(Category) %>%
  summarise(Genes = n_distinct(GeneID), GO_terms = n_distinct(GO_ID),
            Associations = n(), .groups = "drop") %>%
  mutate(Category = recode(Category, "Function" = "Molecular Function (MF)",
                           "Process" = "Biological Process (BP)",
                           "Component" = "Cellular Component (CC)")) %>%
  arrange(desc(Associations)) %>%
  datatable(caption = "GO annotation coverage by ontology", rownames = FALSE,
            options = list(dom = "t", pageLength = 3))

# Evidence-code distribution (IEA usually dominates for non-model organisms)
gene2go_sp %>%
  count(Evidence, sort = TRUE) %>%
  mutate(Evidence = fct_reorder(Evidence, n)) %>%
  ggplot(aes(Evidence, n)) +
  geom_col(fill = "#3C8DAD") + coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(title = paste0("GO evidence codes — ", SPECIES_NAME),
       x = NULL, y = "Number of associations") +
  theme_minimal(base_size = 13)
```

## Load symbol lookup (optional)

```{r load-gene-info}
if (file.exists(GENE_INFO_FILE)) {
  gene_info_sp  <- read.delim(GENE_INFO_FILE, stringsAsFactors = FALSE)
  HAS_GENE_INFO <- TRUE
  cat(sprintf("Symbol lookup loaded: %s genes.\n", comma(nrow(gene_info_sp))))
} else {
  HAS_GENE_INFO <- FALSE
  message("Symbol lookup not found (", GENE_INFO_FILE, "). ",
          "Only numeric NCBI Gene IDs will be accepted as input.")
}
```

## DE gene list and background

The input format is auto-detected: all-numeric entries are treated as NCBI Gene
IDs; otherwise they are treated as symbols and converted via the lookup table.

```{r input-gene-list}
read_gene_lines <- function(path) {
  x <- readLines(path)
  x <- x[!grepl("^\\s*#", x)]           # drop comment lines
  x <- trimws(sub("#.*", "", x))        # strip inline comments
  x[nchar(x) > 0]
}

to_ids <- function(values, what = "input") {
  if (length(values) == 0) stop("No entries found in ", what, ".")
  if (all(grepl("^\\d+$", values))) {
    cat(what, "detected as NCBI Gene IDs (numeric).\n")
    return(as.integer(values))
  }
  cat(what, "detected as gene symbols.\n")
  if (!HAS_GENE_INFO)
    stop("Symbols detected in ", what, " but no lookup table is available. ",
         "Provide ", GENE_INFO_FILE, " or use numeric IDs.")
  convert_symbols_to_ids(values, gene_info_sp)$ids
}

# ── DE genes ──
if (!file.exists(DE_GENE_FILE))
  stop("DE gene list not found: ", DE_GENE_FILE)
de_gene_ids <- to_ids(read_gene_lines(DE_GENE_FILE), "DE list")
cat("DE gene IDs read:", length(de_gene_ids), "\n")

# ── Background ──
if (!is.null(BACKGROUND_FILE) && file.exists(BACKGROUND_FILE)) {
  background_gene_ids <- to_ids(read_gene_lines(BACKGROUND_FILE), "Background")
  cat("Custom background:", length(background_gene_ids), "genes\n")
} else {
  background_gene_ids <- unique(gene2go_sp$GeneID)
  cat("Background = all annotated genes:", length(background_gene_ids), "genes\n")
}

# ── Keep only DE genes present in the background ──
de_not_bg   <- setdiff(de_gene_ids, background_gene_ids)
de_gene_ids <- intersect(de_gene_ids, background_gene_ids)
cat(sprintf("  %d DE genes retained (in background); %d dropped (no annotation)\n",
            length(de_gene_ids), length(de_not_bg)))
if (length(de_gene_ids) == 0)
  stop("None of the DE genes are present in the background — cannot run enrichment.")

# Character vectors used by clusterProfiler throughout
de_chars <- as.character(de_gene_ids)
bg_chars <- as.character(background_gene_ids)
```

---

# GO enrichment

## Build term-to-gene mappings

`enricher()` needs a `TERM2GENE` (`term, gene`) and `TERM2NAME` (`term, name`)
table. We build one pair per ontology so BP / MF / CC are tested separately.

```{r prepare-mappings}
category_map <- c("Process" = "BP", "Function" = "MF", "Component" = "CC")

ontology_data <- gene2go_sp %>%
  mutate(Ontology = category_map[Category]) %>%
  filter(!is.na(Ontology)) %>%
  split(.$Ontology)

t2g_list <- map(ontology_data, ~ .x %>%
                  transmute(term = GO_ID, gene = as.character(GeneID)) %>% distinct())
t2n_list <- map(ontology_data, ~ .x %>%
                  transmute(term = GO_ID, name = GO_term) %>% distinct())

cat("Per-ontology TERM2GENE sizes:\n")
walk2(names(t2g_list), t2g_list, ~ cat(sprintf("  %s: %s associations, %s terms\n",
      .x, comma(nrow(.y)), comma(n_distinct(.y$term)))))
```

## Run `enricher()` (BP / MF / CC)

```{r run-enrichment}
enrich_results <- imap(t2g_list, function(t2g, ont) {
  enricher(
    gene          = de_chars,
    universe      = bg_chars,
    TERM2GENE     = t2g,
    TERM2NAME     = t2n_list[[ont]],
    pvalueCutoff  = PVALUE_CUTOFF,
    qvalueCutoff  = QVALUE_CUTOFF,
    pAdjustMethod = PADJ_METHOD,
    minGSSize     = MIN_GENE_SET,
    maxGSSize     = MAX_GENE_SET
  )
})

# Enforce the ≥ MIN_DE_IN_TERM overlap requirement on every ontology
enrich_results <- map(enrich_results, filter_min_overlap)

cat(sprintf(
  "Filters applied: ≥ %d DE gene(s)/term, background set size in [%d, %d], padj = %s.\n\n",
  MIN_DE_IN_TERM, MIN_GENE_SET, MAX_GENE_SET, PADJ_METHOD))

# Per-ontology summary
imap_dfr(enrich_results, function(res, ont) tibble(
  Ontology                       = ont,
  `DE genes annotated`           = length(intersect(de_chars, t2g_list[[ont]]$gene)),
  `Background genes annotated`   = length(intersect(bg_chars, t2g_list[[ont]]$gene)),
  `Terms reported`               = if (has_rows(res)) nrow(as.data.frame(res)) else 0L,
  `Significant (padj < cutoff)`  = n_sig(res)
)) %>%
  datatable(rownames = FALSE, options = list(dom = "t", pageLength = 3),
            caption = "GO enrichment summary per ontology")
```

## Result tables {.tabset}

### BP
```{r bp-table}
enrich_table(enrich_results[["BP"]], "Enriched Biological Process GO terms")
```

### MF
```{r mf-table}
enrich_table(enrich_results[["MF"]], "Enriched Molecular Function GO terms")
```

### CC
```{r cc-table}
enrich_table(enrich_results[["CC"]], "Enriched Cellular Component GO terms")
```

## Tidy results table

A single tidy data frame (fold enrichment, `-log10` padj) reused by the plots.

```{r combined-df}
combined_df <- imap_dfr(enrich_results, function(res, ont) {
  if (!has_rows(res)) return(NULL)
  as.data.frame(res) %>%
    mutate(Ontology       = ont,
           GeneRatioNum   = parse_ratio(GeneRatio),
           BgRatioNum     = parse_ratio(BgRatio),
           FoldEnrichment = GeneRatioNum / BgRatioNum,
           neg_log10_padj = -log10(p.adjust))
})

if (nrow(combined_df) == 0)
  combined_df <- tibble(ID = character(), Description = character(),
                        GeneRatio = character(), BgRatio = character(),
                        pvalue = numeric(), p.adjust = numeric(), qvalue = numeric(),
                        geneID = character(), Count = integer(), Ontology = character(),
                        GeneRatioNum = numeric(), BgRatioNum = numeric(),
                        FoldEnrichment = numeric(), neg_log10_padj = numeric())

cat(sprintf("Combined GO table: %d terms across %d ontologies.\n",
            nrow(combined_df), n_distinct(combined_df$Ontology)))
```

## Visualisations {.tabset}

A curated, non-redundant set: dot plot, bar plot, enrichment map, gene-concept
network and gene–term heatmap per ontology, followed by a combined overview.

### Dot plots
```{r go-dotplots, fig.height=8}
for (ont in names(enrich_results)) {
  res <- enrich_results[[ont]]
  if (has_rows(res)) print(enrich_dotplot(res, paste0("GO ", ont, " — Dot plot")))
}
```

### Bar plots
```{r go-barplots, fig.height=8}
for (ont in names(enrich_results)) {
  res <- enrich_results[[ont]]
  if (has_rows(res)) print(enrich_barplot(res, title = paste0("GO ", ont, " — Bar plot")))
}
```

### Enrichment maps
```{r go-emap, fig.width=12, fig.height=10}
for (ont in names(enrich_results)) {
  res <- enrich_results[[ont]]
  if (has_rows(res, 2)) tryCatch(
    print(emapplot(pairwise_termsim(res), showCategory = 30) +
            ggtitle(paste0("GO ", ont, " — Enrichment map"))),
    error = function(e) message("Emap skipped for ", ont, ": ", e$message))
}
```

### Gene-concept networks
```{r go-cnet, fig.width=13, fig.height=10}
for (ont in names(enrich_results)) {
  res <- enrich_results[[ont]]
  if (has_rows(res, 2)) tryCatch(
    print(cnetplot(res, showCategory = 10) +
            ggtitle(paste0("GO ", ont, " — Gene-concept network"))),
    error = function(e) message("Cnetplot skipped for ", ont, ": ", e$message))
}
```

### Gene–term heatmaps
```{r go-heatplot, fig.width=14, fig.height=9}
for (ont in names(enrich_results)) {
  res <- enrich_results[[ont]]
  if (has_rows(res, 2)) tryCatch(
    print(heatplot(res, showCategory = SHOW_CATEGORY) +
            ggtitle(paste0("GO ", ont, " — Gene–term heatmap")) +
            theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))),
    error = function(e) message("Heatplot skipped for ", ont, ": ", e$message))
}
```

### Combined overview
```{r go-combined, fig.width=13, fig.height=9}
if (nrow(combined_df) > 0) {
  # Top terms per ontology, one bar chart
  bar_df <- combined_df %>% group_by(Ontology) %>% slice_min(p.adjust, n = 10) %>%
    ungroup() %>% mutate(Description = str_wrap(Description, 45),
                         Description = fct_reorder(Description, neg_log10_padj))
  print(
    ggplot(bar_df, aes(neg_log10_padj, Description, fill = Ontology)) +
      geom_col(width = 0.7) +
      scale_fill_manual(values = ontology_colours) +
      geom_vline(xintercept = -log10(PVALUE_CUTOFF), linetype = "dashed", colour = "grey40") +
      labs(title = paste0("Top enriched GO terms — ", SPECIES_NAME),
           subtitle = paste0("Dashed line = padj ", PVALUE_CUTOFF),
           x = expression(-log[10](p[adj])), y = NULL) +
      theme_minimal(base_size = 13) +
      theme(plot.title = element_text(face = "bold"), legend.position = "top")
  )

  # Enrichment volcano (fold enrichment vs significance)
  print(
    ggplot(combined_df, aes(log2(FoldEnrichment), neg_log10_padj)) +
      geom_point(aes(size = Count, colour = Ontology), alpha = 0.7) +
      scale_colour_manual(values = ontology_colours) +
      scale_size_continuous(range = c(2, 9), name = "Gene count") +
      geom_hline(yintercept = -log10(PVALUE_CUTOFF), linetype = "dashed", colour = "grey40") +
      geom_vline(xintercept = 0, linetype = "dotted", colour = "grey60") +
      ggrepel::geom_text_repel(
        data = combined_df %>% group_by(Ontology) %>% slice_min(p.adjust, n = 3) %>% ungroup(),
        aes(label = str_trunc(Description, 35)), size = 3, max.overlaps = 15,
        box.padding = 0.5, show.legend = FALSE) +
      labs(title = "GO enrichment volcano", x = expression(log[2](Fold~Enrichment)),
           y = expression(-log[10](p[adj]))) +
      theme_minimal(base_size = 12) +
      theme(plot.title = element_text(face = "bold", size = 14))
  )
}
```

## DAG-aware enrichment (topGO)

`enricher()` treats GO terms independently. `topGO`'s `weight01` algorithm
walks the GO hierarchy and down-weights broad parent terms whose children are
already significant, giving more specific, less redundant results.

Three raw p-value columns are reported per term: **`classic`** (a plain Fisher
test on DAG-propagated counts, no hierarchy weighting), **`weight01`** and
**`elim`** (both DAG-aware). Only `classic` is FDR-adjusted, and it is shown
**two ways, side by side**, because the choice of BH denominator changes the
result substantially:

- **`classic_padj_ora`** — BH over *testable* terms only (≥ 1 DE gene and
  background size ≤ `MAX_GENE_SET`). This matches the `enricher()`/ORA convention
  used elsewhere in this pipeline and is the value to compare against `enricher`.
- **`classic_padj_allGO`** — BH over *every* scored node in the ontology (the
  strict "whole-ontology scan" FDR). Thousands of terms have no DE genes and
  p ≈ 1, so they inflate the denominator; this column is markedly more
  conservative and is provided for reference / stricter reporting.

`weight01` and `elim` are intentionally left as raw p-values: their algorithms
already account for the GO DAG structure and should not be BH-corrected again
(Alexa & Rahnenführer, 2009).

```{r topgo-run}
topgo_results <- list()
if (RUN_TOPGO && !requireNamespace("topGO", quietly = TRUE)) {
  message("topGO not installed — skipping DAG-aware enrichment.")
  RUN_TOPGO <- FALSE
}

if (RUN_TOPGO) {
  suppressPackageStartupMessages(library(topGO))

  gene2GO_list <- split(gene2go_sp$GO_ID, as.character(gene2go_sp$GeneID))
  all_genes    <- factor(as.integer(bg_chars %in% de_chars)); names(all_genes) <- bg_chars

  for (ont in c("BP", "MF", "CC")) {
    GOdata <- tryCatch(
      new("topGOdata", ontology = ont, allGenes = all_genes,
          annot = annFUN.gene2GO, gene2GO = gene2GO_list, nodeSize = MIN_GENE_SET),
      error = function(e) { message("topGOdata failed for ", ont, ": ", e$message); NULL })
    if (is.null(GOdata)) next

    res_classic  <- runTest(GOdata, algorithm = "classic",  statistic = "fisher")
    res_weight01 <- runTest(GOdata, algorithm = "weight01", statistic = "fisher")
    res_elim     <- runTest(GOdata, algorithm = "elim",     statistic = "fisher")

    # ── FDR for the classic algorithm — reported TWO ways, side by side ─────
    # "classic" scores EVERY ontology node that survived nodeSize pruning,
    # including thousands with zero DE genes (Fisher p ≈ 1). The denominator of
    # the BH correction changes the result substantially, so both conventions
    # are reported:
    #   • classic_padj_ora   : BH over "testable" terms only (≥ 1 DE gene AND
    #                          background ≤ MAX_GENE_SET) — matches enricher()/ORA.
    #   • classic_padj_allGO : BH over EVERY scored ontology node — the strict
    #                          "whole-ontology scan" FDR (much more conservative).
    # Both use the FULL term set (not just the top rows GenTable prints).
    # weight01/elim stay as raw p-values: their algorithms already account for
    # the GO DAG and should not be BH-corrected again (Alexa & Rahnenführer 2009).
    ts        <- termStat(GOdata)          # Annotated / Significant / Expected per GO.ID
    classic_p <- score(res_classic)        # raw Fisher p-value per GO.ID
    testable  <- rownames(ts)[ts$Significant >= 1 & ts$Annotated <= MAX_GENE_SET]
    padj_ora  <- setNames(rep(NA_real_, length(classic_p)), names(classic_p))
    padj_ora[testable] <- p.adjust(classic_p[testable], method = PADJ_METHOD)
    padj_df   <- tibble(GO.ID              = names(classic_p),
                        classic_padj_ora   = padj_ora,
                        classic_padj_allGO = p.adjust(classic_p, method = PADJ_METHOD))

    allRes <- GenTable(GOdata, classic = res_classic, weight01 = res_weight01,
                       elim = res_elim, orderBy = "weight01", ranksOf = "classic",
                       topNodes = min(50, length(usedGO(GOdata)))) %>%
      mutate(across(c(classic, weight01, elim), ~ as.numeric(gsub("< ", "", .x))),
             Ontology = ont) %>%
      left_join(padj_df, by = "GO.ID") %>%
      filter(Significant >= MIN_DE_IN_TERM)     # ≥ MIN_DE_IN_TERM DE genes per term

    topgo_results[[ont]] <- allRes
    message("  topGO ", ont, ": ",
            sum(allRes$weight01 < PVALUE_CUTOFF, na.rm = TRUE),
            " terms with weight01 p < ", PVALUE_CUTOFF)
  }
}
topgo_combined <- if (length(topgo_results)) bind_rows(topgo_results) else tibble()

# Renderer for a single ontology's topGO table (used in the tabs below)
topgo_table <- function(ont) {
  res <- topgo_results[[ont]]
  if (is.null(res) || nrow(res) == 0) {
    cat("topGO results not available for", ont, "\n"); return(invisible())
  }
  res %>%
    mutate(across(c(classic, weight01, elim, classic_padj_ora, classic_padj_allGO),
                  ~ signif(.x, 3))) %>%
    datatable(rownames = FALSE,
              caption = paste0("topGO ", ont, " — raw p (classic/weight01/elim) plus ",
                               "classic FDR two ways: _ora (testable terms) vs _allGO (whole ontology)"),
              options = list(pageLength = 15, scrollX = TRUE))
}
```

### topGO results {.tabset}

#### BP
```{r topgo-bp}
topgo_table("BP")
```

#### MF
```{r topgo-mf}
topgo_table("MF")
```

#### CC
```{r topgo-cc}
topgo_table("CC")
```

### topGO plots

```{r topgo-plots, fig.width=12, fig.height=8}
if (nrow(topgo_combined) > 0) {
  # weight01 vs classic — points below the diagonal gain significance after DAG correction
  print(
    ggplot(topgo_combined, aes(-log10(classic), -log10(weight01))) +
      geom_point(aes(colour = Ontology, size = Significant), alpha = 0.6) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
      scale_colour_manual(values = ontology_colours) +
      scale_size_continuous(range = c(1.5, 7), name = "DE genes\nin term") +
      ggrepel::geom_text_repel(
        data = topgo_combined %>% group_by(Ontology) %>% slice_min(weight01, n = 5) %>% ungroup(),
        aes(label = str_trunc(Term, 35)), size = 3, max.overlaps = 15, show.legend = FALSE) +
      labs(title = "topGO: classic vs weight01",
           subtitle = "Below diagonal = more significant after DAG correction",
           x = expression(-log[10](p[classic])), y = expression(-log[10](p[weight01]))) +
      theme_minimal(base_size = 12) + theme(plot.title = element_text(face = "bold", size = 14))
  )

  topgo_bar <- topgo_combined %>% filter(weight01 < PVALUE_CUTOFF) %>%
    group_by(Ontology) %>% slice_min(weight01, n = 10) %>% ungroup() %>%
    mutate(Term = str_wrap(Term, 45), Term = fct_reorder(Term, -log10(weight01)))
  if (nrow(topgo_bar) > 0)
    print(
      ggplot(topgo_bar, aes(-log10(weight01), Term, fill = Ontology)) +
        geom_col(width = 0.7) + scale_fill_manual(values = ontology_colours) +
        geom_vline(xintercept = -log10(PVALUE_CUTOFF), linetype = "dashed", colour = "grey40") +
        labs(title = "topGO weight01 — top enriched terms",
             x = expression(-log[10](p[weight01])), y = NULL) +
        theme_minimal(base_size = 12) +
        theme(plot.title = element_text(face = "bold", size = 14), legend.position = "top")
    )
}
```

## Export GO results

```{r go-export}
# Tables
imap(enrich_results, function(res, ont) {
  if (has_rows(res)) write.table(as.data.frame(res),
      file.path(OUT_DIR, paste0("GO_enrichment_", ont, ".tsv")),
      sep = "\t", row.names = FALSE, quote = FALSE)
})
if (nrow(combined_df) > 0)
  write.table(combined_df, file.path(OUT_DIR, "GO_enrichment_combined.tsv"),
              sep = "\t", row.names = FALSE, quote = FALSE)
for (ont in names(topgo_results))
  write.table(topgo_results[[ont]], file.path(OUT_DIR, paste0("topGO_", ont, ".tsv")),
              sep = "\t", row.names = FALSE, quote = FALSE)

# Core plots
for (ont in names(enrich_results)) {
  res <- enrich_results[[ont]]; if (!has_rows(res)) next
  safe_save(paste0("dotplot_", ont, ".png"), function() enrich_dotplot(res, paste0("GO ", ont)))
  safe_save(paste0("barplot_", ont, ".png"), function() enrich_barplot(res, title = paste0("GO ", ont)))
  if (has_rows(res, 2)) {
    safe_save(paste0("emap_", ont, ".png"),
              function() emapplot(pairwise_termsim(res), showCategory = 30), w = 12, h = 10)
    safe_save(paste0("cnetplot_", ont, ".png"),
              function() cnetplot(res, showCategory = 10), w = 13, h = 10)
  }
}
cat("GO tables and plots written to:", OUT_DIR, "\n")
```

---

# Pathway enrichment

KEGG can run **directly** when your species is a registered KEGG organism
(here `phyp`). Reactome, WikiPathways and PANTHER cover only model organisms, so
those go through a **reference organism** via orthologs (here zebrafish). The
reference and all database codes are set in the parameters; the ortholog map is
built once and cached as `ORTHO_FILE`. To retarget a different species/reference
(e.g. a bird → chicken) see **[Adapting to another species](#adapting)**.

> Ordering note: ortholog mapping runs **before** KEGG so that KEGG can also use
> orthologs (via `KEGG_VIA_ORTHOLOGS`) when your species is absent from KEGG.

## Pathway packages

```{r pathway-packages}
pkg_ok <- function(p) requireNamespace(p, quietly = TRUE)
# ORTHO_ORGDB is loaded by name so the reference organism can be swapped from
# the parameters (org.Dr.eg.db, org.Gg.eg.db, org.Hs.eg.db, ...).
for (p in c("ReactomePA", "DOSE", ORTHO_ORGDB, "biomaRt", "KEGGREST"))
  if (pkg_ok(p)) suppressPackageStartupMessages(library(p, character.only = TRUE))

HAS_KEGG    <- pkg_ok("KEGGREST")
HAS_REACT   <- pkg_ok("ReactomePA")
HAS_BIOMART <- pkg_ok("biomaRt")
HAS_ORGDB   <- pkg_ok(ORTHO_ORGDB)                 # reference-organism annotation package
ORG_DB      <- if (HAS_ORGDB) get(ORTHO_ORGDB) else NULL   # the AnnotationDbi object itself
HAS_RBIOAPI <- pkg_ok("rbioapi")
if (HAS_RBIOAPI) library(rbioapi)

RUN_ORTHO_PATHWAYS <- RUN_PATHWAY_ENRICHMENT && HAS_BIOMART && HAS_ORGDB
```

## Ortholog mapping to the reference organism

Maps your species' NCBI Gene IDs → the **reference organism's** Entrez IDs,
first via Ensembl BioMart, then falling back to gene-symbol matching against
`ORTHO_ORGDB`. Driven entirely by the ortholog parameters at the top — no code
edits needed to change the reference (see **[Adapting to another species](#adapting)**).
The result is cached to `ORTHO_FILE`; delete that file to rebuild.

```{r ortholog-mapping}
de_ref_entrez <- character(); bg_ref_entrez <- character(); ortho_map <- NULL
ENOUGH_ORTHOLOGS <- FALSE

if (RUN_ORTHO_PATHWAYS) {
  if (file.exists(ORTHO_FILE)) {
    ortho_map <- read.delim(ORTHO_FILE, stringsAsFactors = FALSE)
    cat("Loaded cached ortholog map:", comma(nrow(ortho_map)), "rows\n")
  } else {
    # Homolog attribute / dataset names for the chosen reference organism
    attr_symbol  <- paste0(ORTHO_BIOMART_PREFIX, "_homolog_associated_gene_name")
    attr_ensembl <- paste0(ORTHO_BIOMART_PREFIX, "_homolog_ensembl_gene")

    # Strategy 1: Ensembl BioMart species-to-species orthology
    ortho_map <- tryCatch({
      if (nchar(SPECIES_BIOMART_REGEX) == 0) stop("BioMart disabled (empty SPECIES_BIOMART_REGEX)")
      ensembl <- useMart("ensembl")
      sp_hit  <- listDatasets(ensembl) %>%
        filter(grepl(SPECIES_BIOMART_REGEX, dataset, ignore.case = TRUE))
      if (nrow(sp_hit) == 0) stop("your species is not in Ensembl")
      sp_mart <- useDataset(sp_hit$dataset[1], mart = ensembl)
      bm <- getBM(attributes = c("entrezgene_id", attr_symbol, attr_ensembl), mart = sp_mart)
      colnames(bm) <- c("sp_geneid", "ref_symbol", "ref_ensembl")
      bm <- bm %>% filter(!is.na(sp_geneid), ref_ensembl != "")
      ref_mart <- useDataset(ORTHO_ENSEMBL_DATASET, mart = ensembl)
      ref_ids  <- getBM(attributes = c("ensembl_gene_id", "entrezgene_id"),
                        filters = "ensembl_gene_id", values = unique(bm$ref_ensembl),
                        mart = ref_mart) %>%
        filter(!is.na(entrezgene_id)) %>%
        dplyr::rename(ref_ensembl = ensembl_gene_id, ref_entrez = entrezgene_id)
      bm %>% left_join(ref_ids, by = "ref_ensembl") %>% filter(!is.na(ref_entrez)) %>%
        distinct(sp_geneid, ref_entrez, .keep_all = TRUE)
    }, error = function(e) { message("BioMart unavailable (", e$message,
                                     ") — using symbol matching."); NULL })

    # Strategy 2 (fallback): gene-symbol matching against ORTHO_ORGDB
    if (is.null(ortho_map) || nrow(ortho_map) == 0) {
      if (HAS_GENE_INFO && HAS_ORGDB) {
        sp_sym <- gene_info_sp %>% dplyr::select(GeneID, Symbol) %>%
          filter(Symbol != "-") %>% mutate(sym_lower = tolower(Symbol))
        ref_sym <- AnnotationDbi::select(ORG_DB, keys = keys(ORG_DB, "SYMBOL"),
                     columns = c("SYMBOL", "ENTREZID"), keytype = "SYMBOL") %>%
          filter(!is.na(ENTREZID)) %>% mutate(sym_lower = tolower(SYMBOL))
        ortho_map <- sp_sym %>%
          inner_join(ref_sym, by = "sym_lower", relationship = "many-to-many") %>%
          transmute(sp_geneid = GeneID, ref_symbol = SYMBOL,
                    ref_ensembl = NA_character_, ref_entrez = as.integer(ENTREZID)) %>%
          distinct(sp_geneid, ref_entrez, .keep_all = TRUE)
      } else ortho_map <- tibble(sp_geneid = integer(), ref_symbol = character(),
                                 ref_ensembl = character(), ref_entrez = integer())
    }
    if (nrow(ortho_map) > 0)
      write.table(ortho_map, ORTHO_FILE, sep = "\t", row.names = FALSE, quote = FALSE)
  }

  # A cached/user-supplied ORTHO_FILE may use legacy column names; normalise them.
  if (!is.null(ortho_map) && nrow(ortho_map) > 0) {
    if (!"ref_entrez" %in% colnames(ortho_map) && "dr_entrez" %in% colnames(ortho_map))
      ortho_map <- dplyr::rename(ortho_map, ref_entrez = dr_entrez)
    if (!"sp_geneid" %in% colnames(ortho_map) && "GeneID1" %in% colnames(ortho_map))
      ortho_map <- dplyr::rename(ortho_map, sp_geneid = GeneID1)
    stopifnot(all(c("sp_geneid", "ref_entrez") %in% colnames(ortho_map)))
  }

  de_ortho      <- ortho_map %>% filter(sp_geneid %in% de_gene_ids)
  bg_ortho      <- ortho_map %>% filter(sp_geneid %in% background_gene_ids)
  de_ref_entrez <- unique(as.character(de_ortho$ref_entrez))
  bg_ref_entrez <- unique(as.character(bg_ortho$ref_entrez))

  cat(sprintf("Orthologs → %s: %s/%s DE genes → %s ref IDs; %s/%s background → %s ref IDs\n",
      ORTHO_ORGDB,
      comma(n_distinct(de_ortho$sp_geneid)), comma(length(de_gene_ids)), comma(length(de_ref_entrez)),
      comma(n_distinct(bg_ortho$sp_geneid)), comma(length(background_gene_ids)), comma(length(bg_ref_entrez))))

  ENOUGH_ORTHOLOGS <- length(de_ref_entrez) >= MIN_GENE_SET
  if (!ENOUGH_ORTHOLOGS)
    message("Few DE orthologs (", length(de_ref_entrez),
            ") — Reactome / WikiPathways / PANTHER may return nothing.")
} else {
  message("Ortholog mapping skipped (needs biomaRt + ", ORTHO_ORGDB,
          ", or pathway analysis is off).")
}
```

## KEGG {.tabset}

Runs directly on your NCBI IDs when `KEGG_VIA_ORTHOLOGS = FALSE`, or on the
reference-organism orthologs (with `KEGG_ORG` set to that organism's code, e.g.
`gga`) when `TRUE`.

### Results
```{r kegg-run}
kegg_result <- NULL
kegg_genes  <- if (KEGG_VIA_ORTHOLOGS) de_ref_entrez else de_chars
kegg_univ   <- if (KEGG_VIA_ORTHOLOGS) bg_ref_entrez else bg_chars
kegg_ready  <- RUN_PATHWAY_ENRICHMENT && HAS_KEGG &&
               (!KEGG_VIA_ORTHOLOGS || (ENOUGH_ORTHOLOGS && length(kegg_genes) > 0))

if (kegg_ready) {
  kegg_result <- tryCatch(
    enrichKEGG(gene = kegg_genes, universe = kegg_univ, organism = KEGG_ORG,
               keyType = "ncbi-geneid", pvalueCutoff = PVALUE_CUTOFF,
               qvalueCutoff = QVALUE_CUTOFF, pAdjustMethod = PADJ_METHOD,
               minGSSize = MIN_GENE_SET, maxGSSize = MAX_GENE_SET,
               use_internal_data = FALSE),
    error = function(e) { message("KEGG failed (API?): ", e$message); NULL })
  kegg_result <- filter_min_overlap(kegg_result)
  cat(sprintf("KEGG (%s, %s): %d significant pathway(s).\n", KEGG_ORG,
              if (KEGG_VIA_ORTHOLOGS) "orthologs" else "direct", n_sig(kegg_result)))
} else cat("KEGG enrichment skipped.\n")

enrich_table(kegg_result, paste0("KEGG pathway enrichment (", KEGG_ORG, ")"))
```

### Plots
```{r kegg-plots, fig.width=12, fig.height=8}
if (has_rows(kegg_result)) {
  print(enrich_dotplot(kegg_result, "KEGG — Dot plot"))
  print(enrich_barplot(kegg_result, title = "KEGG — Bar plot"))
  if (has_rows(kegg_result, 2)) {
    tryCatch(print(emapplot(pairwise_termsim(kegg_result), showCategory = 25) +
                     ggtitle("KEGG — Enrichment map")),
             error = function(e) message("KEGG emap skipped: ", e$message))
    tryCatch(print(cnetplot(kegg_result, showCategory = 8) +
                     ggtitle("KEGG — Gene-concept network")),
             error = function(e) message("KEGG cnetplot skipped: ", e$message))
  }
}
```

## Reactome (via orthologs) {.tabset}

### Results
```{r reactome-run}
reactome_result <- NULL
if (RUN_ORTHO_PATHWAYS && ENOUGH_ORTHOLOGS && HAS_REACT) {
  reactome_result <- tryCatch(
    ReactomePA::enrichPathway(gene = de_ref_entrez, universe = bg_ref_entrez,
      organism = REACTOME_ORG, pvalueCutoff = PVALUE_CUTOFF, qvalueCutoff = QVALUE_CUTOFF,
      pAdjustMethod = PADJ_METHOD, minGSSize = MIN_GENE_SET, maxGSSize = MAX_GENE_SET,
      readable = TRUE),
    error = function(e) { message("Reactome failed: ", e$message); NULL })
  reactome_result <- filter_min_overlap(reactome_result)
  cat(sprintf("Reactome: %d significant pathway(s).\n", n_sig(reactome_result)))
} else cat("Reactome enrichment skipped.\n")

enrich_table(reactome_result, "Reactome pathway enrichment (via zebrafish orthologs)")
```

### Plots
```{r reactome-plots, fig.width=12, fig.height=8}
if (has_rows(reactome_result)) {
  print(enrich_dotplot(reactome_result, "Reactome — Dot plot"))
  print(enrich_barplot(reactome_result, title = "Reactome — Bar plot"))
  if (has_rows(reactome_result, 2))
    tryCatch(print(emapplot(pairwise_termsim(reactome_result), showCategory = 25) +
                     ggtitle("Reactome — Enrichment map")),
             error = function(e) message("Reactome emap skipped: ", e$message))
}
```

## WikiPathways (via orthologs) {.tabset}

### Results
```{r wp-run}
wp_result <- NULL
if (RUN_ORTHO_PATHWAYS && ENOUGH_ORTHOLOGS) {
  wp_result <- tryCatch(
    enrichWP(gene = de_ref_entrez, universe = bg_ref_entrez, organism = WP_ORGANISM,
             pvalueCutoff = PVALUE_CUTOFF, qvalueCutoff = QVALUE_CUTOFF,
             pAdjustMethod = PADJ_METHOD, minGSSize = MIN_GENE_SET, maxGSSize = MAX_GENE_SET),
    error = function(e) { message("WikiPathways failed: ", e$message); NULL })
  wp_result <- filter_min_overlap(wp_result)
  cat(sprintf("WikiPathways: %d significant pathway(s).\n", n_sig(wp_result)))
} else cat("WikiPathways enrichment skipped.\n")

enrich_table(wp_result, "WikiPathways enrichment (via zebrafish orthologs)")
```

### Plots
```{r wp-plots, fig.width=12, fig.height=8}
if (has_rows(wp_result)) {
  print(enrich_dotplot(wp_result, "WikiPathways — Dot plot"))
  print(enrich_barplot(wp_result, title = "WikiPathways — Bar plot"))
  if (has_rows(wp_result, 2))
    tryCatch(print(emapplot(pairwise_termsim(wp_result), showCategory = 20) +
                     ggtitle("WikiPathways — Enrichment map")),
             error = function(e) message("WP emap skipped: ", e$message))
}
```

## PANTHER (via orthologs, PANTHER API) {.tabset}

PANTHER is not built into `clusterProfiler`; we query its own overrepresentation
API via `rbioapi`. Note PANTHER uses its **own whole-genome zebrafish reference**
as background, so its p-values are not strictly comparable to the ortholog-
background KEGG/Reactome/WikiPathways results above.

### Results
```{r panther-run}
panther_clean <- tibble()
if (RUN_ORTHO_PATHWAYS && ENOUGH_ORTHOLOGS && HAS_RBIOAPI) {
  panther_raw <- tryCatch({
    r <- rba_panther_enrich(genes = de_ref_entrez, organism = PANTHER_TAXID,
           annot_dataset = "ANNOT_TYPE_ID_PANTHER_PATHWAY",
           test_type = "FISHER", correction = "FDR", cutoff = 1)
    if (is.data.frame(r$result)) r$result else if (is.data.frame(r)) r else NULL
  }, error = function(e) { message("PANTHER API failed: ", e$message); NULL })

  if (!is.null(panther_raw) && nrow(panther_raw) > 0) {
    panther_clean <- panther_raw %>%
      filter(!is.na(term.id), term.id != "UNCLASSIFIED") %>%
      mutate(pvalue = as.numeric(pValue), fdr = as.numeric(fdr),
             across(c(number_in_list, expected, number_in_reference, fold_enrichment), as.numeric)) %>%
      filter(!is.na(pvalue),
             number_in_list >= MIN_DE_IN_TERM,        # ≥ MIN_DE_IN_TERM DE genes
             number_in_reference >= MIN_GENE_SET,     # ≥ MIN_GENE_SET in reference
             number_in_reference <= MAX_GENE_SET) %>%
      mutate(fdr = pmax(fdr, .Machine$double.xmin)) %>%
      arrange(fdr)
    cat(sprintf("PANTHER: %d significant pathway(s) (FDR < %s).\n",
                sum(panther_clean$fdr < PVALUE_CUTOFF, na.rm = TRUE), PVALUE_CUTOFF))
  } else cat("PANTHER returned no usable results.\n")
} else cat("PANTHER enrichment skipped (needs rbioapi + orthologs).\n")

if (nrow(panther_clean) > 0)
  panther_clean %>%
    dplyr::select(ID = term.id, Description = term.label, Genes_in_list = number_in_list,
                  Genes_in_ref = number_in_reference, Fold_Enrichment = fold_enrichment,
                  pvalue, FDR = fdr) %>%
    mutate(across(c(pvalue, FDR, Fold_Enrichment), ~ signif(.x, 3))) %>%
    datatable(rownames = FALSE, caption = "PANTHER pathway enrichment (zebrafish orthologs)",
              options = list(pageLength = 15, scrollX = TRUE)) else cat("No PANTHER table.\n")
```

### Plot
```{r panther-plot, fig.width=11, fig.height=8}
panther_sig <- panther_clean %>% filter(fdr < PVALUE_CUTOFF) %>% slice_min(fdr, n = 20)
if (nrow(panther_sig) > 0) {
  panther_sig <- panther_sig %>%
    mutate(neg_log10_fdr = -log10(fdr), term.label = str_wrap(term.label, 50),
           term.label = fct_reorder(term.label, neg_log10_fdr))
  print(
    ggplot(panther_sig, aes(neg_log10_fdr, term.label)) +
      geom_col(aes(fill = fold_enrichment), width = 0.7) +
      scale_fill_gradient(low = "#FDD49E", high = "#B30000", name = "Fold enrichment") +
      geom_vline(xintercept = -log10(PVALUE_CUTOFF), linetype = "dashed", colour = "grey40") +
      labs(title = "PANTHER — Bar plot", x = expression(-log[10](FDR)), y = NULL) +
      theme_minimal(base_size = 12) + theme(plot.title = element_text(face = "bold", size = 14))
  )
}
```

## Combined pathway overview {.tabset}

```{r pathway-combined-build}
# One tidy frame across all pathway databases
all_pathway_df <- tibble()
for (db in list(list(n = "KEGG", r = kegg_result),
                list(n = "Reactome", r = reactome_result),
                list(n = "WikiPathways", r = wp_result))) {
  if (has_rows(db$r))
    all_pathway_df <- bind_rows(all_pathway_df, as.data.frame(db$r) %>%
      mutate(Database = db$n, GeneRatioNum = parse_ratio(GeneRatio),
             BgRatioNum = parse_ratio(BgRatio),
             FoldEnrichment = GeneRatioNum / BgRatioNum, neg_log10_padj = -log10(p.adjust)))
}
if (nrow(panther_clean) > 0)
  all_pathway_df <- bind_rows(all_pathway_df, panther_clean %>%
    transmute(ID = term.id, Description = term.label, Count = number_in_list,
              p.adjust = fdr, pvalue = pvalue, Database = "PANTHER",
              GeneRatioNum = number_in_list / length(de_ref_entrez),
              BgRatioNum = number_in_reference / max(length(bg_ref_entrez), 1),
              FoldEnrichment = fold_enrichment, neg_log10_padj = -log10(fdr)))
```

### Cross-database summary
```{r pathway-summary-table}
bind_rows(
  tibble(Database = c("KEGG", "Reactome", "WikiPathways"),
         Strategy = c("Direct (phyp)", "Zebrafish orthologs", "Zebrafish orthologs"),
         Significant = c(n_sig(kegg_result), n_sig(reactome_result), n_sig(wp_result))),
  tibble(Database = "PANTHER", Strategy = "Zebrafish orthologs (API)",
         Significant = sum(panther_clean$fdr < PVALUE_CUTOFF, na.rm = TRUE))
) %>%
  datatable(rownames = FALSE, caption = "Cross-database pathway enrichment summary",
            options = list(dom = "t", pageLength = 4))
```

### Combined plots
```{r pathway-combined-plots, fig.width=13, fig.height=9}
if (nrow(all_pathway_df) > 0) {
  plot_df <- all_pathway_df %>% group_by(Database) %>% slice_min(p.adjust, n = 8) %>%
    ungroup() %>% mutate(Description = str_wrap(Description, 50),
                         Description = fct_reorder(Description, neg_log10_padj))
  print(
    ggplot(plot_df, aes(GeneRatioNum, Description)) +
      geom_point(aes(size = Count, colour = neg_log10_padj, shape = Database)) +
      scale_colour_viridis_c(option = "inferno", direction = -1,
                             name = expression(-log[10](p[adj]))) +
      scale_shape_manual(values = c(KEGG = 16, Reactome = 17, WikiPathways = 15, PANTHER = 18)) +
      scale_size_continuous(range = c(3, 10), name = "Gene count") +
      labs(title = paste0("Pathway enrichment overview — ", SPECIES_NAME),
           subtitle = "KEGG = direct; others = via zebrafish orthologs",
           x = "Gene ratio (k / n)", y = NULL) +
      theme_minimal(base_size = 12) + theme(plot.title = element_text(face = "bold", size = 14))
  )
  print(
    ggplot(all_pathway_df, aes(log2(FoldEnrichment), neg_log10_padj)) +
      geom_point(aes(size = Count, colour = Database), alpha = 0.7) +
      scale_colour_manual(values = pathway_colours) +
      scale_size_continuous(range = c(2, 9), name = "Gene count") +
      geom_hline(yintercept = -log10(PVALUE_CUTOFF), linetype = "dashed", colour = "grey40") +
      geom_vline(xintercept = 0, linetype = "dotted", colour = "grey60") +
      ggrepel::geom_text_repel(
        data = all_pathway_df %>% group_by(Database) %>% slice_min(p.adjust, n = 2) %>% ungroup(),
        aes(label = str_trunc(Description, 40)), size = 3, max.overlaps = 15, show.legend = FALSE) +
      labs(title = "Pathway enrichment volcano — all databases",
           x = expression(log[2](Fold~Enrichment)), y = expression(-log[10](p[adj]))) +
      theme_minimal(base_size = 12) + theme(plot.title = element_text(face = "bold", size = 14))
  )
} else cat("No pathway results to combine.\n")
```

## Export pathway results

```{r pathway-export}
for (db in list(list(n = "KEGG", r = kegg_result),
                list(n = "Reactome", r = reactome_result),
                list(n = "WikiPathways", r = wp_result)))
  if (has_rows(db$r))
    write.table(as.data.frame(db$r),
                file.path(PATHWAY_DIR, paste0(db$n, "_enrichment.tsv")),
                sep = "\t", row.names = FALSE, quote = FALSE)
if (nrow(panther_clean) > 0)
  write.table(panther_clean, file.path(PATHWAY_DIR, "PANTHER_enrichment.tsv"),
              sep = "\t", row.names = FALSE, quote = FALSE)
if (nrow(all_pathway_df) > 0)
  write.table(all_pathway_df, file.path(PATHWAY_DIR, "pathway_enrichment_combined.tsv"),
              sep = "\t", row.names = FALSE, quote = FALSE)
if (!is.null(ortho_map) && nrow(ortho_map) > 0)
  write.table(ortho_map, file.path(PATHWAY_DIR, "ortholog_mapping_used.tsv"),
              sep = "\t", row.names = FALSE, quote = FALSE)

# Core pathway plots
for (db in list(list(n = "KEGG", r = kegg_result),
                list(n = "Reactome", r = reactome_result),
                list(n = "WikiPathways", r = wp_result))) {
  if (!has_rows(db$r)) next
  local({ r <- db$r; nm <- db$n
    safe_save(paste0("dotplot_", nm, ".png"), function() enrich_dotplot(r, nm),
              dir = PATHWAY_PLOT_DIR)
    safe_save(paste0("barplot_", nm, ".png"), function() enrich_barplot(r, title = nm),
              dir = PATHWAY_PLOT_DIR)
  })
}
cat("Pathway tables and plots written to:", PATHWAY_DIR, "and", PATHWAY_PLOT_DIR, "\n")
```

---

# Validation & benchmarking

## Manual hypergeometric test vs `enricher()` {.tabset}

For transparency we reproduce the ORA by hand with the one-sided hypergeometric
test (identical to `fisher.test(alternative = "greater")`, but vectorised and
numerically robust). Restricting each term's gene set to the universe — exactly
as `enricher()` does — is what keeps the test valid under a custom background.

```{r manual-ora-fn}
manual_ora <- function(t2g, de, universe,
                       min_de = MIN_DE_IN_TERM, min_size = MIN_GENE_SET,
                       max_size = MAX_GENE_SET, padj_method = PADJ_METHOD) {
  uni  <- intersect(universe, unique(t2g$gene))   # universe genes annotated in this ontology
  de_u <- intersect(de, uni)
  N <- length(uni); n <- length(de_u)
  t2g %>%
    filter(gene %in% uni) %>%                      # restrict K to the universe
    group_by(term) %>%
    summarise(K = n_distinct(gene), k = sum(gene %in% de_u), .groups = "drop") %>%
    filter(K >= min_size, K <= max_size, k >= 1) %>%
    mutate(pvalue = phyper(k - 1, K, N - K, n, lower.tail = FALSE),
           fold_enrichment = (k / n) / (K / N),
           GeneRatio = paste0(k, "/", n), BgRatio = paste0(K, "/", N),
           padj = p.adjust(pvalue, method = padj_method)) %>%   # BH over all tested terms
    filter(k >= min_de) %>%                                     # then overlap filter
    arrange(padj)
}

manual_results <- imap(t2g_list, ~ manual_ora(.x, de_chars, bg_chars) %>%
                         left_join(t2n_list[[.y]], by = "term") %>% mutate(Ontology = .y))
```

### BP
```{r manual-bp}
manual_results[["BP"]] %>% filter(padj < PVALUE_CUTOFF) %>%
  dplyr::select(term, name, k, K, GeneRatio, BgRatio, fold_enrichment, pvalue, padj) %>%
  mutate(across(c(pvalue, padj), ~ signif(.x, 3)), fold_enrichment = round(fold_enrichment, 2)) %>%
  datatable(rownames = FALSE, caption = "Manual hypergeometric — BP (should match enricher BP)",
            options = list(pageLength = 10, scrollX = TRUE))
```

### MF
```{r manual-mf}
manual_results[["MF"]] %>% filter(padj < PVALUE_CUTOFF) %>%
  dplyr::select(term, name, k, K, GeneRatio, BgRatio, fold_enrichment, pvalue, padj) %>%
  mutate(across(c(pvalue, padj), ~ signif(.x, 3)), fold_enrichment = round(fold_enrichment, 2)) %>%
  datatable(rownames = FALSE, caption = "Manual hypergeometric — MF (should match enricher MF)",
            options = list(pageLength = 10, scrollX = TRUE))
```

### CC
```{r manual-cc}
manual_results[["CC"]] %>% filter(padj < PVALUE_CUTOFF) %>%
  dplyr::select(term, name, k, K, GeneRatio, BgRatio, fold_enrichment, pvalue, padj) %>%
  mutate(across(c(pvalue, padj), ~ signif(.x, 3)), fold_enrichment = round(fold_enrichment, 2)) %>%
  datatable(rownames = FALSE, caption = "Manual hypergeometric — CC (should match enricher CC)",
            options = list(pageLength = 10, scrollX = TRUE))
```

### Agreement with `enricher()`
```{r manual-vs-enricher, fig.width=9, fig.height=7}
manual_all <- bind_rows(manual_results) %>% dplyr::select(ID = term, Ontology, manual_padj = padj)
check_df <- combined_df %>% dplyr::select(ID, Ontology, enricher_padj = p.adjust) %>%
  inner_join(manual_all, by = c("ID", "Ontology"))

if (nrow(check_df) >= 3) {
  r <- cor(-log10(check_df$enricher_padj), -log10(check_df$manual_padj))
  print(
    ggplot(check_df, aes(-log10(enricher_padj), -log10(manual_padj))) +
      geom_point(aes(colour = Ontology), alpha = 0.7, size = 2.5) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
      scale_colour_manual(values = ontology_colours) +
      labs(title = sprintf("Manual hypergeometric vs enricher (r = %.3f)", r),
           subtitle = "Points on the diagonal confirm the two implementations agree",
           x = expression(-log[10](enricher~p[adj])),
           y = expression(-log[10](manual~p[adj]))) +
      theme_minimal(base_size = 12) + theme(plot.title = element_text(face = "bold", size = 14))
  )
} else cat("Too few shared terms to plot agreement.\n")
```

## `enricher()` vs `topGO` {.tabset}

Terms found by **both** methods are high-confidence; terms unique to `topGO` are
usually more specific (hierarchy-aware). Thresholds differ by design — enricher
uses BH-adjusted p; topGO `weight01` uses its raw DAG-aware p (which should not
be BH-corrected further; Alexa & Rahnenführer, 2009) — so this is a qualitative
overlap, not a like-for-like test.

```{r enricher-vs-topgo-build}
comparison_df <- tibble()
if (length(topgo_results) > 0 && nrow(combined_df) > 0) {
  comparison_df <- imap_dfr(topgo_results, function(tg, ont) {
    en <- combined_df %>% filter(Ontology == ont, p.adjust < PVALUE_CUTOFF) %>%
      dplyr::select(GO_ID = ID, Description, enricher_padj = p.adjust)
    tg2 <- tg %>% filter(weight01 < PVALUE_CUTOFF) %>%
      dplyr::select(GO_ID = GO.ID, Term, topGO_weight01 = weight01)
    full_join(en, tg2, by = "GO_ID") %>%
      mutate(Ontology = ont, Name = coalesce(Description, Term),
             Found_in = case_when(!is.na(enricher_padj) & !is.na(topGO_weight01) ~ "Both",
                                  !is.na(enricher_padj) ~ "enricher only",
                                  TRUE ~ "topGO only")) %>%
      dplyr::select(Ontology, GO_ID, Name, Found_in, enricher_padj, topGO_weight01)
  })
}
```

### Overlap table
```{r comparison-table}
if (nrow(comparison_df) > 0)
  comparison_df %>% mutate(across(c(enricher_padj, topGO_weight01), ~ signif(.x, 3))) %>%
    arrange(Found_in, Ontology) %>%
    datatable(rownames = FALSE, filter = "top",
              caption = "enricher() vs topGO weight01 — term overlap",
              options = list(pageLength = 20, scrollX = TRUE)) else cat("Comparison not available.\n")
```

### Overlap counts
```{r comparison-plot, fig.width=10, fig.height=6}
if (nrow(comparison_df) > 0)
  print(
    comparison_df %>% count(Ontology, Found_in, name = "n") %>%
      ggplot(aes(Ontology, n, fill = Found_in)) +
      geom_col(position = "dodge", width = 0.7) +
      geom_text(aes(label = n), position = position_dodge(0.7), vjust = -0.3, fontface = "bold") +
      scale_fill_manual(values = c(Both = "#2C7BB6", `enricher only` = "#D7191C",
                                   `topGO only` = "#1A9641"), name = "Found in") +
      labs(title = "enricher() vs topGO — term overlap by ontology",
           x = NULL, y = "Significant terms") +
      theme_minimal(base_size = 13) +
      theme(plot.title = element_text(face = "bold", size = 14), legend.position = "top")
  )
```

### Significance concordance
For terms found by **both** methods, how well do the two p-values agree? Note
the axes use different correction schemes by design — enricher is BH-adjusted,
topGO `weight01` is a raw DAG-aware p — so the dashed diagonal is a reference
line, not an expectation of equality.

```{r comparison-concordance, fig.width=10, fig.height=8}
both_df <- if (nrow(comparison_df) > 0 && "Found_in" %in% names(comparison_df)) {
  comparison_df %>%
    filter(Found_in == "Both", !is.na(enricher_padj), !is.na(topGO_weight01))
} else comparison_df[0, ]

if (nrow(both_df) >= 3) {
  print(
    ggplot(both_df, aes(-log10(enricher_padj), -log10(topGO_weight01))) +
      geom_point(aes(colour = Ontology), size = 3, alpha = 0.7) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
      scale_colour_manual(values = ontology_colours) +
      ggrepel::geom_text_repel(
        data = both_df %>% slice_min(enricher_padj, n = 8),
        aes(label = str_trunc(Name, 35)), size = 3, max.overlaps = 12, show.legend = FALSE) +
      labs(title = "Significance concordance — enricher() vs topGO",
           subtitle = "Diagonal shown for reference (axes use different correction methods)",
           x = expression(-log[10](enricher~p[adj]~(BH))),
           y = expression(-log[10](topGO~weight01~p~(raw)))) +
      theme_minimal(base_size = 12) +
      theme(plot.title = element_text(face = "bold", size = 14))
  )
} else {
  cat("Fewer than 3 terms found by both methods — skipping concordance plot.\n")
}
```

---

# Summary

## Parameters used

```{r summary-params}
tibble(Parameter = c("Species (TaxID)", "p-value cutoff", "q-value cutoff",
                     "Adjust method", "Min DE genes / term", "Min background / term",
                     "Max background / term", "topGO", "Pathway enrichment", "Background"),
       Value = c(sprintf("%s (%d)", SPECIES_NAME, TAXID),
                 PVALUE_CUTOFF, QVALUE_CUTOFF, PADJ_METHOD,
                 MIN_DE_IN_TERM, MIN_GENE_SET, MAX_GENE_SET,
                 ifelse(length(topgo_results) > 0, "run", "off/skipped"),
                 ifelse(RUN_PATHWAY_ENRICHMENT, "run", "off"),
                 ifelse(is.null(BACKGROUND_FILE), "all annotated genes", BACKGROUND_FILE)) ) %>%
  datatable(rownames = FALSE, caption = "Parameters used in this run",
            options = list(dom = "t", pageLength = 10))
```

## Enrichment counts across all methods

```{r summary-counts}
summary_counts <- bind_rows(
  imap_dfr(enrich_results, ~ tibble(Analysis = "GO enricher", Set = .y, Significant = n_sig(.x))),
  if (length(topgo_results) > 0)
    imap_dfr(topgo_results, ~ tibble(Analysis = "GO topGO (weight01)", Set = .y,
             Significant = sum(.x$weight01 < PVALUE_CUTOFF, na.rm = TRUE))) else NULL,
  tibble(Analysis = "Pathway", Set = "KEGG",         Significant = n_sig(kegg_result)),
  tibble(Analysis = "Pathway", Set = "Reactome",     Significant = n_sig(reactome_result)),
  tibble(Analysis = "Pathway", Set = "WikiPathways", Significant = n_sig(wp_result)),
  tibble(Analysis = "Pathway", Set = "PANTHER",
         Significant = sum(panther_clean$fdr < PVALUE_CUTOFF, na.rm = TRUE))
)

datatable(summary_counts, rownames = FALSE,
          caption = sprintf("Significant terms/pathways per method (padj/FDR < %s)", PVALUE_CUTOFF),
          options = list(dom = "t", pageLength = 15))
```

## Integrated overview

Top hits from every GO ontology and pathway database in one chart.

```{r integrated-summary, fig.width=15, fig.height=12}
integrated_df <- bind_rows(
  if (nrow(combined_df) > 0)
    combined_df %>% mutate(Database = paste0("GO-", Ontology)) %>%
      group_by(Database) %>% slice_min(p.adjust, n = 5) %>% ungroup() %>%
      dplyr::select(Description, neg_log10_padj, Database) else NULL,
  if (nrow(all_pathway_df) > 0)
    all_pathway_df %>% group_by(Database) %>% slice_min(p.adjust, n = 5) %>% ungroup() %>%
      dplyr::select(Description, neg_log10_padj, Database) else NULL
)

if (nrow(integrated_df) > 0) {
  integrated_df <- integrated_df %>%
    mutate(Description = str_wrap(Description, 45),
           Description = fct_reorder(Description, neg_log10_padj))
  all_colours <- c(`GO-BP` = "#E64B35", `GO-MF` = "#4DBBD5", `GO-CC` = "#00A087",
                   KEGG = "#E64B35BB", Reactome = "#4DBBD5BB",
                   WikiPathways = "#00A087BB", PANTHER = "#F39B7F")
  p <- ggplot(integrated_df, aes(neg_log10_padj, Description, fill = Database)) +
    geom_col(width = 0.7) +
    scale_fill_manual(values = all_colours, name = "Source") +
    geom_vline(xintercept = -log10(PVALUE_CUTOFF), linetype = "dashed", colour = "grey40") +
    labs(title = paste0("Integrated enrichment summary — ", SPECIES_NAME),
         subtitle = "Top 5 per GO ontology and per pathway database",
         x = expression(-log[10](p[adj])), y = NULL) +
    theme_minimal(base_size = 12) +
    theme(plot.title = element_text(face = "bold", size = 14), legend.position = "top") +
    guides(fill = guide_legend(nrow = 1))
  print(p)
  safe_save("integrated_summary.png", function() p, w = 15, h = 12)
} else cat("No enrichment results to integrate.\n")

cat("\nAll outputs saved under:", normalizePath(OUT_DIR), "\n")
```

# Session info

```{r session-info}
sessionInfo()
```
