---
title: "GO Enrichment Analysis for Non-Model Organisms"
subtitle: "*Pangasianodon hypophthalmus* (Striped Catfish) — Using `clusterProfiler::enricher()`"
author: "Phuong Nguyen"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 3
    number_sections: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 10,
  fig.height = 7,
  fig.align  = "center",
  cache      = TRUE
)
```

# Overview

This pipeline performs Gene Ontology (GO) enrichment analysis for a non-model
organism (*Pangasianodon hypophthalmus*, striped catfish, NCBI TaxID: **310915**)
using NCBI's `gene2go` annotation file and the `clusterProfiler::enricher()`
function, which accepts any custom gene-to-term mapping.

**Workflow summary:**

1. Load a pre-filtered organism GO annotation file (`Phypophthalmus_gene2go.tsv`).
2. Read the DE gene list from an external text file (`de_gene_list.txt`).
3. Run over-representation analysis (ORA) with `enricher()` for each GO
   ontology (BP, MF, CC).
4. Visualise results with 15+ plot types: dot plots, bar plots, enrichment
   maps, tree plots, category–gene networks, heatmaps, upset plots, lollipop
   plots, bubble plots, GO circle plots, waterfall plots, enrichment volcanos,
   word clouds, faceted dot matrices, and publication-ready combined figures.
5. Export all results as TSV and all plots as 300-dpi PNGs.

**Required input files** (place in the working directory):

| File | Description |
|------|-------------|
| `Phypophthalmus_gene2go.tsv` | Organism GO annotations (see preparation section) |
| `de_gene_list.txt` | Your DE genes — NCBI Gene IDs **or** gene symbols, one per line |
| `Phypophthalmus_gene_info.tsv` | *(Optional)* Symbol ↔ ID lookup. Required only if your gene list uses symbols instead of numeric IDs |

# Setup

## Install / Load Packages

```{r install-packages, eval=FALSE}
# ── Run once ──
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install(c("clusterProfiler", "enrichplot", "GO.db",
                       "AnnotationDbi", "topGO"))
install.packages(c("tidyverse", "DT", "scales", "ggrepel", "ggwordcloud"))
```

```{r load-packages}
library(tidyverse)
library(clusterProfiler)
library(enrichplot)
library(GO.db)
library(AnnotationDbi)
library(DT)
library(scales)
library(ggrepel)
# ggwordcloud loaded conditionally in the word cloud chunk

# ── Resolve namespace conflicts ──
# AnnotationDbi::select() and IRanges::slice() mask dplyr equivalents
select <- dplyr::select
slice  <- dplyr::slice
```

## Define Parameters

```{r params}
# ── Organism ──
TAXID        <- 310915
SPECIES_NAME <- "Pangasianodon hypophthalmus"

# ── Enrichment thresholds ──
PVALUE_CUTOFF  <- 0.05
QVALUE_CUTOFF  <- 0.2
MIN_GENE_SET   <- 5
MAX_GENE_SET   <- 500

# ── Colour palette for ontology facets ──
ontology_colours <- c("BP" = "#E64B35", "MF" = "#4DBBD5", "CC" = "#00A087")

# ── Output directory ──
OUT_DIR <- "GO_results"
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)
```

# Preparing `gene2go_sp` (One-Time Pre-Processing)

The `gene2go` file from NCBI's Gene FTP maps every Gene ID to its associated GO
terms.  It is \~1.3 GB compressed and covers **all organisms** in NCBI Gene.
You only need to do this **once** — the filtered output file
(`Phypophthalmus_gene2go.tsv`) is small and reusable across analyses.

## Step-by-step instructions

**Step 1.** Download the file from NCBI FTP (terminal or R):

```bash
# Terminal — fastest method
wget https://ftp.ncbi.nlm.nih.gov/gene/DATA/gene2go.gz
gunzip gene2go.gz
```

Or in R:

```{r download-instructions, eval=FALSE}
download.file(
  url      = "https://ftp.ncbi.nlm.nih.gov/gene/DATA/gene2go.gz",
  destfile = "gene2go.gz",
  mode     = "wb"
)
R.utils::gunzip("gene2go.gz", remove = TRUE)
```

**Step 2.** Read the full file and filter for your organism:

```{r filter-instructions, eval=FALSE}
gene2go_raw <- read.delim("gene2go", header = TRUE, comment.char = "#",
                           stringsAsFactors = FALSE)

colnames(gene2go_raw) <- c("TaxID", "GeneID", "GO_ID", "Evidence",
                            "Qualifier", "GO_term", "PubMed", "Category")

# ── Filter for Pangasianodon hypophthalmus (TaxID 310915) ──
# Change the TaxID below for a different organism.
# Common examples:
#   Oreochromis niloticus (Nile tilapia)  = 8128
#   Ictalurus punctatus (channel catfish)  = 7998
#   Crassostrea gigas (Pacific oyster)     = 29159
gene2go_sp <- gene2go_raw[gene2go_raw$TaxID == 310915, ]

write.table(gene2go_sp, file = "Phypophthalmus_gene2go.tsv",
            sep = "\t", row.names = FALSE, quote = FALSE)
```

**Step 3.** You can now delete the large `gene2go` file — only the small
filtered TSV (`Phypophthalmus_gene2go.tsv`) is needed by this pipeline.

---

# Load Pre-Filtered GO Annotations

This pipeline assumes `Phypophthalmus_gene2go.tsv` already exists in the
working directory (see instructions above).

```{r load-gene2go}
GENE2GO_FILE <- "Phypophthalmus_gene2go.tsv"

if (!file.exists(GENE2GO_FILE)) {
  stop("File not found: ", GENE2GO_FILE,
       "\n  Please follow the preparation steps above to generate it first.")
}

gene2go_sp <- read.delim(GENE2GO_FILE, stringsAsFactors = FALSE)

cat(sprintf(
  "GO annotations for %s (TaxID %d):\n  %s unique genes\n  %s unique GO terms\n  %s gene–GO associations\n",
  SPECIES_NAME, TAXID,
  comma(n_distinct(gene2go_sp$GeneID)),
  comma(n_distinct(gene2go_sp$GO_ID)),
  comma(nrow(gene2go_sp))
))
```

## Annotation Summary by Ontology

```{r annotation-summary}
gene2go_sp %>%
  group_by(Category) %>%
  summarise(
    Genes      = n_distinct(GeneID),
    GO_terms   = n_distinct(GO_ID),
    Associations = n(),
    .groups = "drop"
  ) %>%
  mutate(Category = recode(Category,
    "Function"  = "Molecular Function (MF)",
    "Process"   = "Biological Process (BP)",
    "Component" = "Cellular Component (CC)"
  )) %>%
  arrange(desc(Associations)) %>%
  datatable(
    caption  = "GO annotation coverage by ontology",
    rownames = FALSE,
    options  = list(dom = "t", pageLength = 3)
  )
```

## Evidence Code Distribution

```{r evidence-codes, fig.height=5}
gene2go_sp %>%
  count(Evidence, sort = TRUE) %>%
  mutate(Evidence = fct_reorder(Evidence, n)) %>%
  ggplot(aes(x = Evidence, y = n)) +
  geom_col(fill = "#3C8DAD") +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(
    title    = paste0("GO Evidence Code Distribution — ", SPECIES_NAME),
    subtitle = "IEA (Inferred from Electronic Annotation) dominates in non-model organisms",
    x = NULL, y = "Number of associations"
  ) +
  theme_minimal(base_size = 13)
```

# Input Gene List

## Preparing the Symbol ↔ ID Lookup Table (One-Time)

If your DE gene list uses **gene symbols** (e.g., `tlr3`, `nfkb1`) instead of
NCBI Gene IDs, you need a mapping table.  NCBI's `gene_info` file provides
this for all organisms.  Prepare it once, just like `gene2go`:

```bash
# Terminal — download gene_info (~2 GB compressed)
wget https://ftp.ncbi.nlm.nih.gov/gene/DATA/gene_info.gz
gunzip gene_info.gz
```

```{r gene-info-prep-instructions, eval=FALSE}
# ── Alternative: download in R ──
download.file("https://ftp.ncbi.nlm.nih.gov/gene/DATA/gene_info.gz",
              "gene_info.gz", mode = "wb")
R.utils::gunzip("gene_info.gz", remove = TRUE)

# ── Read and filter for your organism ──
# this following steps take forever to run, so we need to use bash to extract gene_info with TaxID we want
#gene_info_raw <- read.delim("gene_info", header = TRUE, comment.char = "#",
                             quote = "", stringsAsFactors = FALSE)
#colnames(gene_info_raw)[1] <- "TaxID"

### Assume that we already filter gene_info for TaxID "310915" (P. hypophthalmus) -> gene_info_sp.txt available

gene_info_sp0 <- read.delim("gene_info_sp.txt", header = TRUE,
                             quote = "", stringsAsFactors = FALSE)
colnames(gene_info_sp0)[1] <- "TaxID"

gene_info_sp <- gene_info_sp0 %>%
  select(GeneID, Symbol, Synonyms, description, type_of_gene)

write.table(gene_info_sp, file = "Phypophthalmus_gene_info.tsv",
            sep = "\t", row.names = FALSE, quote = FALSE)

# You can now delete the large gene_info file
```

The filtered file (`Phypophthalmus_gene_info.tsv`) is small (~3 MB) and maps
every gene symbol (and synonym) to its NCBI Gene ID.

## Load Symbol Lookup Table

```{r load-gene-info}
GENE_INFO_FILE <- "Phypophthalmus_gene_info.tsv"

if (file.exists(GENE_INFO_FILE)) {
  gene_info_sp <- read.delim(GENE_INFO_FILE, stringsAsFactors = FALSE)
  HAS_GENE_INFO <- TRUE
  cat(sprintf("Gene info loaded: %s genes with symbol mappings\n",
              comma(nrow(gene_info_sp))))
} else {
  HAS_GENE_INFO <- FALSE
  message("Gene info file not found: ", GENE_INFO_FILE,
          "\n  Symbol-to-ID conversion will not be available.",
          "\n  Only numeric NCBI Gene IDs will be accepted as input.",
          "\n  See instructions above to generate the lookup table.")
}
```

## Symbol → NCBI Gene ID Converter

```{r symbol-converter}
convert_symbols_to_ids <- function(symbols, gene_info) {
  # ── Step 1: exact match on Symbol column (case-insensitive) ──
  lookup <- gene_info %>%
    mutate(Symbol_lower = tolower(Symbol))

  symbols_lower <- tolower(trimws(symbols))
  match_df <- tibble(input = symbols, input_lower = symbols_lower) %>%
    left_join(lookup %>% dplyr::select(GeneID, Symbol, Symbol_lower),
              by = c("input_lower" = "Symbol_lower"))

  matched   <- match_df %>% filter(!is.na(GeneID))
  unmatched <- match_df %>% filter(is.na(GeneID))

  # ── Step 2: search Synonyms column for remaining symbols ──
  if (nrow(unmatched) > 0 && "Synonyms" %in% colnames(gene_info)) {
    synonym_lookup <- gene_info %>%
      filter(Synonyms != "-") %>%
      mutate(syn_list = strsplit(Synonyms, "\\|")) %>%
      unnest(syn_list) %>%
      mutate(syn_lower = tolower(trimws(syn_list))) %>%
      dplyr::select(GeneID, Symbol, syn_lower)

    synonym_match <- unmatched %>%
      dplyr::select(input, input_lower) %>%
      left_join(synonym_lookup, by = c("input_lower" = "syn_lower"))

    newly_matched <- synonym_match %>% filter(!is.na(GeneID))
    still_unmatched <- synonym_match %>% filter(is.na(GeneID))

    matched <- bind_rows(
      matched %>% dplyr::select(input, GeneID, Symbol),
      newly_matched %>% dplyr::select(input, GeneID, Symbol)
    )
    unmatched <- still_unmatched
  }

  # ── Report ──
  cat(sprintf("Symbol conversion:\n  %d / %d symbols matched to NCBI Gene IDs\n",
              nrow(matched), length(symbols)))

  if (nrow(unmatched) > 0) {
    cat(sprintf("  %d symbols NOT found:\n", nrow(unmatched)))
    cat("    ", paste(head(unmatched$input, 20), collapse = ", "))
    if (nrow(unmatched) > 20) cat(" ...")
    cat("\n")
  }

  # ── Handle duplicates (one symbol mapping to multiple IDs) ──
  dupes <- matched %>% group_by(input) %>% filter(n() > 1)
  if (nrow(dupes) > 0) {
    cat(sprintf("  %d symbols mapped to multiple IDs (first match kept):\n",
                n_distinct(dupes$input)))
    matched <- matched %>% group_by(input) %>% dplyr::slice(1) %>% ungroup()
  }

  list(
    ids       = matched$GeneID,
    matched   = matched,
    unmatched = unmatched$input
  )
}
```

## Read Input Gene List

The pipeline auto-detects your input format:

- **Numeric** → treated as NCBI Gene IDs directly.
- **Non-numeric** → treated as gene symbols and converted via the lookup table.

```{r input-gene-list}
# ══════════════════════════════════════════════════════════════════════
#                    *** EDIT THESE PATHS ***
# ══════════════════════════════════════════════════════════════════════
DE_GENE_FILE     <- "de_gene_list_symbols.txt"     # gene IDs or symbols (one per line)
BACKGROUND_FILE  <- NULL                    # set to a file path for custom background,
                                            # or leave NULL to use all annotated genes
# ══════════════════════════════════════════════════════════════════════

# ── Read DE gene list (raw lines, skip comments) ──
if (!file.exists(DE_GENE_FILE)) {
  stop("DE gene list not found: ", DE_GENE_FILE,
       "\n  Please provide a text file with one gene per line.")
}

raw_lines <- readLines(DE_GENE_FILE)
raw_lines <- raw_lines[!grepl("^\\s*#", raw_lines) & nchar(trimws(raw_lines)) > 0]
raw_values <- trimws(sub("#.*", "", raw_lines))   # strip inline comments
raw_values <- raw_values[nchar(raw_values) > 0]

cat("Entries read from file:", length(raw_values), "\n")

# ── Auto-detect format: numeric IDs vs. symbols ──
is_numeric_input <- all(grepl("^\\d+$", raw_values))

if (is_numeric_input) {
  cat("Input detected as: NCBI Gene IDs (numeric)\n")
  de_gene_ids <- as.integer(raw_values)
} else {
  cat("Input detected as: Gene symbols (non-numeric)\n")
  if (!HAS_GENE_INFO) {
    stop("Gene symbols detected but no lookup table available.\n",
         "  Please generate ", GENE_INFO_FILE, " first (see instructions above),\n",
         "  or provide NCBI Gene IDs instead.")
  }
  conversion <- convert_symbols_to_ids(raw_values, gene_info_sp)
  de_gene_ids <- conversion$ids
}

cat("Gene IDs to test:", length(de_gene_ids), "\n")

# ── Background: custom file or all annotated genes ──
if (!is.null(BACKGROUND_FILE) && file.exists(BACKGROUND_FILE)) {
  bg_raw <- readLines(BACKGROUND_FILE)
  bg_raw <- bg_raw[!grepl("^\\s*#", bg_raw) & nchar(trimws(bg_raw)) > 0]
  bg_values <- trimws(sub("#.*", "", bg_raw))
  bg_is_numeric <- all(grepl("^\\d+$", bg_values))

  if (bg_is_numeric) {
    background_gene_ids <- as.integer(bg_values)
  } else if (HAS_GENE_INFO) {
    bg_conv <- convert_symbols_to_ids(bg_values, gene_info_sp)
    background_gene_ids <- bg_conv$ids
  } else {
    stop("Background file contains symbols but no lookup table is available.")
  }
  cat("Custom background read:", length(background_gene_ids), "genes\n")
} else {
  background_gene_ids <- unique(gene2go_sp$GeneID)
  cat("Background (all annotated genes):", length(background_gene_ids), "genes\n")
}

# ── Retain only DE genes present in the background ──
de_in_bg  <- intersect(de_gene_ids, background_gene_ids)
de_not_bg <- setdiff(de_gene_ids, background_gene_ids)

cat(sprintf(
  "  %d DE genes matched the background\n  %d DE genes had no GO annotation (excluded)\n",
  length(de_in_bg), length(de_not_bg)
))

if (length(de_not_bg) > 0 && length(de_not_bg) <= 20) {
  cat("  Excluded IDs:", paste(de_not_bg, collapse = ", "), "\n")
}

de_gene_ids <- de_in_bg
```

# GO Enrichment with `clusterProfiler::enricher()`

## Prepare TERM2GENE and TERM2NAME Mappings

`enricher()` requires two data frames:

- **TERM2GENE**: columns `[term, gene]` — the full mapping of GO terms to genes.
- **TERM2NAME**: columns `[term, name]` — human-readable names for each term.

```{r prepare-mappings}
# TERM2GENE: GO_ID → GeneID (character)
term2gene <- gene2go_sp %>%
  transmute(term = GO_ID, gene = as.character(GeneID)) %>%
  distinct()

# TERM2NAME: GO_ID → GO_term
term2name <- gene2go_sp %>%
  transmute(term = GO_ID, name = GO_term) %>%
  distinct()

cat(sprintf(
  "TERM2GENE: %s rows  |  TERM2NAME: %s terms\n",
  comma(nrow(term2gene)), comma(nrow(term2name))
))
```

## Split Annotations by Ontology

Running enrichment separately for Biological Process (BP), Molecular Function
(MF), and Cellular Component (CC) makes the results easier to interpret and
avoids mixing ontologies in a single test.

```{r split-ontology}
# Map NCBI categories to short codes
category_map <- c("Process" = "BP", "Function" = "MF", "Component" = "CC")

ontology_data <- gene2go_sp %>%
  mutate(Ontology = category_map[Category]) %>%
  split(.$Ontology)

# Build per-ontology TERM2GENE / TERM2NAME
t2g_list <- map(ontology_data, ~ .x %>%
  transmute(term = GO_ID, gene = as.character(GeneID)) %>%
  distinct()
)

t2n_list <- map(ontology_data, ~ .x %>%
  transmute(term = GO_ID, name = GO_term) %>%
  distinct()
)
```

## Run enricher() for Each Ontology {.tabset}

```{r run-enrichment}
de_chars <- as.character(de_gene_ids)
bg_chars <- as.character(background_gene_ids)

enrich_results <- imap(t2g_list, function(t2g, ont) {
  enricher(
    gene          = de_chars,
    universe      = bg_chars,
    TERM2GENE     = t2g,
    TERM2NAME     = t2n_list[[ont]],
    pvalueCutoff  = PVALUE_CUTOFF,
    qvalueCutoff  = QVALUE_CUTOFF,
    pAdjustMethod = "BH",
    minGSSize     = MIN_GENE_SET,
    maxGSSize     = MAX_GENE_SET
  )
})

# Summary
imap_dfr(enrich_results, function(res, ont) {
  n_sig <- if (!is.null(res) && nrow(res@result %>% filter(p.adjust < PVALUE_CUTOFF)) > 0)
    nrow(res@result %>% filter(p.adjust < PVALUE_CUTOFF)) else 0
  tibble(Ontology = ont, `Significant terms (padj < 0.05)` = n_sig)
}) %>%
  datatable(rownames = FALSE, options = list(dom = "t", pageLength = 3),
            caption = "Number of significantly enriched GO terms per ontology")
```

### BP — Biological Process

```{r bp-table}
if (!is.null(enrich_results[["BP"]]) &&
    nrow(as.data.frame(enrich_results[["BP"]])) > 0) {
  as.data.frame(enrich_results[["BP"]]) %>%
    dplyr::select(ID, Description, GeneRatio, BgRatio, pvalue, p.adjust, qvalue, Count) %>%
    mutate(across(c(pvalue, p.adjust, qvalue), ~ signif(.x, 3))) %>%
    datatable(
      rownames = FALSE,
      caption  = "Enriched Biological Process GO terms",
      options  = list(pageLength = 15, scrollX = TRUE)
    )
} else {
  cat("No significantly enriched BP terms at padj <", PVALUE_CUTOFF, "\n")
}
```

### MF — Molecular Function

```{r mf-table}
if (!is.null(enrich_results[["MF"]]) &&
    nrow(as.data.frame(enrich_results[["MF"]])) > 0) {
  as.data.frame(enrich_results[["MF"]]) %>%
    dplyr::select(ID, Description, GeneRatio, BgRatio, pvalue, p.adjust, qvalue, Count) %>%
    mutate(across(c(pvalue, p.adjust, qvalue), ~ signif(.x, 3))) %>%
    datatable(
      rownames = FALSE,
      caption  = "Enriched Molecular Function GO terms",
      options  = list(pageLength = 15, scrollX = TRUE)
    )
} else {
  cat("No significantly enriched MF terms at padj <", PVALUE_CUTOFF, "\n")
}
```

### CC — Cellular Component

```{r cc-table}
if (!is.null(enrich_results[["CC"]]) &&
    nrow(as.data.frame(enrich_results[["CC"]])) > 0) {
  as.data.frame(enrich_results[["CC"]]) %>%
    dplyr::select(ID, Description, GeneRatio, BgRatio, pvalue, p.adjust, qvalue, Count) %>%
    mutate(across(c(pvalue, p.adjust, qvalue), ~ signif(.x, 3))) %>%
    datatable(
      rownames = FALSE,
      caption  = "Enriched Cellular Component GO terms",
      options  = list(pageLength = 15, scrollX = TRUE)
    )
} else {
  cat("No significantly enriched CC terms at padj <", PVALUE_CUTOFF, "\n")
}
```

## {-}

# Helper: Extract Enrichment Data Frames

Before plotting, we prepare tidy data frames from the enrichment results with
parsed ratios and fold-enrichment values that are reused across multiple plot
types.

```{r helper-extract-df}
# ── Parse GeneRatio "k/n" into numeric ──
parse_ratio <- function(ratio_str) {
  parts <- str_split(ratio_str, "/", simplify = TRUE)
  as.numeric(parts[, 1]) / as.numeric(parts[, 2])
}

# ── Build a combined tidy data frame across all ontologies ──
combined_df <- imap_dfr(enrich_results, function(res, ont) {
  if (!is.null(res) && nrow(as.data.frame(res)) > 0) {
    as.data.frame(res) %>%
      mutate(
        Ontology       = ont,
        GeneRatioNum   = parse_ratio(GeneRatio),
        BgRatioNum     = parse_ratio(BgRatio),
        FoldEnrichment = GeneRatioNum / BgRatioNum,
        neg_log10_padj = -log10(p.adjust),
        RichFactor     = GeneRatioNum  # alias used in some plots
      )
  }
})

cat(sprintf("Combined enrichment table: %d terms across %d ontologies\n",
            nrow(combined_df), n_distinct(combined_df$Ontology)))
```

# Visualisation — enrichplot Built-in Plots {.tabset}

## Dot Plots

Dot plots show enrichment significance on the x-axis and gene ratio (proportion
of DE genes in the term) by dot size.  Colour encodes adjusted p-value.

```{r dotplots, fig.height=8}
for (ont in names(enrich_results)) {
  res <- enrich_results[[ont]]
  if (!is.null(res) && nrow(as.data.frame(res)) > 0) {
    p <- dotplot(res, showCategory = 20, font.size = 11) +
      ggtitle(paste0("GO ", ont, " — Dot Plot")) +
      theme(plot.title = element_text(face = "bold", size = 14))
    print(p)
  }
}
```

## Bar Plots

Classic horizontal bar chart coloured by adjusted p-value.

```{r barplots, fig.height=8}
# Custom bar plot — replaces enrichplot::barplot() which was removed in
# recent enrichplot versions.
enrich_barplot <- function(res, showCategory = 20, font.size = 11) {
  df <- as.data.frame(res) %>%
    head(showCategory) %>%
    mutate(
      GeneRatioNum = sapply(GeneRatio, function(x) {
        parts <- as.numeric(strsplit(x, "/")[[1]])
        parts[1] / parts[2]
      }),
      Description = factor(Description, levels = rev(Description))
    )
  ggplot(df, aes(x = Count, y = Description, fill = p.adjust)) +
    geom_col(width = 0.7) +
    scale_fill_gradient(low = "#E64B35", high = "#4292C6",
                        name = "p.adjust",
                        trans = "log10") +
    labs(x = "Count", y = NULL) +
    theme_minimal(base_size = font.size)
}

for (ont in names(enrich_results)) {
  res <- enrich_results[[ont]]
  if (!is.null(res) && nrow(as.data.frame(res)) > 0) {
    p <- enrich_barplot(res, showCategory = 20, font.size = 11) +
      ggtitle(paste0("GO ", ont, " — Bar Plot")) +
      theme(plot.title = element_text(face = "bold", size = 14))
    print(p)
  }
}
```

## Enrichment Map (Network)

The enrichment map connects GO terms that share overlapping gene sets (Jaccard
similarity).  Clusters of connected nodes highlight functionally related
processes.

```{r emap, fig.width=12, fig.height=10}
for (ont in names(enrich_results)) {
  res <- enrich_results[[ont]]
  if (!is.null(res) && nrow(as.data.frame(res)) >= 2) {
    res_pw <- pairwise_termsim(res)
    p <- emapplot(res_pw, showCategory = 30) +
      ggtitle(paste0("GO ", ont, " — Enrichment Map")) +
      theme(plot.title = element_text(face = "bold", size = 14))
    print(p)
  }
}
```

## Category–Gene Network (cnetplot)

Shows the linkages between GO terms and the actual DE genes annotated to each.
Shared genes between terms appear as connections bridging multiple category
nodes.

```{r cnet, fig.width=13, fig.height=10}
for (ont in names(enrich_results)) {
  res <- enrich_results[[ont]]
  if (!is.null(res) && nrow(as.data.frame(res)) >= 2) {
    tryCatch({
      p <- cnetplot(res, showCategory = 10) +
        ggtitle(paste0("GO ", ont, " — Category–Gene Network")) +
        theme(plot.title = element_text(face = "bold", size = 14))
      print(p)
    }, error = function(e) {
      message("Cnetplot skipped for ", ont, ": ", e$message)
    })
  }
}
```

## Category–Gene Network (Top 5 terms)

A second cnetplot focused on fewer terms for a cleaner view of gene–term
linkages.

```{r cnet-top5, fig.width=13, fig.height=10}
for (ont in names(enrich_results)) {
  res <- enrich_results[[ont]]
  if (!is.null(res) && nrow(as.data.frame(res)) >= 2) {
    tryCatch({
      p <- cnetplot(res, showCategory = 5) +
        ggtitle(paste0("GO ", ont, " — Category–Gene Network (Top 5)")) +
        theme(plot.title = element_text(face = "bold", size = 14))
      print(p)
    }, error = function(e) {
      message("Cnetplot (top 5) skipped for ", ont, ": ", e$message)
    })
  }
}
```

## Tree-Style Clustered Bar Plot

Enriched terms clustered by pairwise gene-set overlap (Jaccard similarity) and
displayed as a bar chart ordered by the dendrogram.  This is a version-safe
replacement for `treeplot()`.

```{r treeplot-custom, fig.width=14, fig.height=10}
for (ont in names(enrich_results)) {
  res <- enrich_results[[ont]]
  if (is.null(res) || nrow(as.data.frame(res)) < 3) next

  tryCatch({
    res_df <- as.data.frame(res) %>% head(25)

    # Build a binary gene × term matrix
    gene_lists <- strsplit(res_df$geneID, "/")
    all_genes  <- unique(unlist(gene_lists))
    mat <- matrix(0, nrow = nrow(res_df), ncol = length(all_genes),
                  dimnames = list(res_df$Description, all_genes))
    for (i in seq_along(gene_lists)) {
      mat[i, gene_lists[[i]]] <- 1
    }

    # Hierarchical clustering on Jaccard distance
    jdist <- dist(mat, method = "binary")
    hc    <- hclust(jdist, method = "ward.D2")

    # Order descriptions by dendrogram
    res_df$Description <- factor(res_df$Description,
                                  levels = res_df$Description[hc$order])

    p <- ggplot(res_df, aes(x = -log10(p.adjust), y = Description,
                             fill = -log10(p.adjust))) +
      geom_col(width = 0.7) +
      scale_fill_gradient(low = "#FDD49E", high = "#B30000",
                          name = expression(-log[10](p[adj]))) +
      labs(
        title = paste0("GO ", ont,
                       " — Clustered Enrichment (Ward.D2 on Jaccard)"),
        x     = expression(-log[10](p[adj])),
        y     = NULL
      ) +
      theme_minimal(base_size = 11) +
      theme(plot.title = element_text(face = "bold", size = 14))
    print(p)
  }, error = function(e) {
    message("Clustered bar plot skipped for ", ont, ": ", e$message)
  })
}
```

## Upset Plot

Upset plots show set intersections — which GO terms share the same DE genes.
Useful for spotting genes that drive multiple enriched terms.

```{r upset, fig.width=12, fig.height=7}
for (ont in names(enrich_results)) {
  res <- enrich_results[[ont]]
  if (!is.null(res) && nrow(as.data.frame(res)) >= 2) {
    tryCatch({
      p <- upsetplot(res, n = 12)
      print(p)
    }, error = function(e) {
      message("Upset plot skipped for ", ont, ": ", e$message)
    })
  }
}
```

## Heatmap-like Plot

A tile/heatmap showing which DE genes are annotated to each enriched GO term.

```{r heatplot, fig.width=14, fig.height=9}
for (ont in names(enrich_results)) {
  res <- enrich_results[[ont]]
  if (!is.null(res) && nrow(as.data.frame(res)) >= 2) {
    tryCatch({
      p <- heatplot(res, showCategory = 20) +
        ggtitle(paste0("GO ", ont, " — Gene–Term Heatmap")) +
        theme(
          plot.title = element_text(face = "bold", size = 14),
          axis.text.x = element_text(angle = 45, hjust = 1, size = 8)
        )
      print(p)
    }, error = function(e) {
      message("Heatplot skipped for ", ont, ": ", e$message)
    })
  }
}
```

## {-}

# Visualisation — Custom ggplot2 Plots {.tabset}

The following plots are built with ggplot2 from the enrichment data frame,
giving full control over aesthetics.  They are commonly used in publications and
supplement the built-in enrichplot functions.

## Lollipop Plot

A cleaner alternative to bar charts.  The dot encodes gene count and the
segment length encodes significance.

```{r lollipop, fig.width=11, fig.height=8}
for (ont in names(enrich_results)) {
  df_ont <- combined_df %>% filter(Ontology == ont) %>%
    slice_min(p.adjust, n = 25) %>%
    mutate(Description = str_wrap(Description, width = 50),
           Description = fct_reorder(Description, neg_log10_padj))

  if (nrow(df_ont) == 0) next

  p <- ggplot(df_ont, aes(x = neg_log10_padj, y = Description)) +
    geom_segment(aes(x = 0, xend = neg_log10_padj, yend = Description),
                 colour = "grey60", linewidth = 0.5) +
    geom_point(aes(size = Count, colour = FoldEnrichment)) +
    scale_colour_gradient(low = "#FDD49E", high = "#D7301F",
                          name = "Fold Enrichment") +
    scale_size_continuous(range = c(2, 8), name = "Gene Count") +
    geom_vline(xintercept = -log10(0.05), linetype = "dashed",
               colour = "grey40", linewidth = 0.4) +
    labs(
      title = paste0("GO ", ont, " — Lollipop Plot"),
      x     = expression(-log[10](p.adjust)),
      y     = NULL
    ) +
    theme_minimal(base_size = 12) +
    theme(plot.title = element_text(face = "bold", size = 14))
  print(p)
}
```

## Bubble Plot (Fold Enrichment × Count)

Bubble plots map fold enrichment to the x-axis, significance to colour, and
gene count to bubble size — encoding three dimensions simultaneously.

```{r bubble, fig.width=12, fig.height=8}
for (ont in names(enrich_results)) {
  df_ont <- combined_df %>% filter(Ontology == ont) %>%
    slice_min(p.adjust, n = 25) %>%
    mutate(Description = str_wrap(Description, width = 45),
           Description = fct_reorder(Description, FoldEnrichment))

  if (nrow(df_ont) == 0) next

  p <- ggplot(df_ont, aes(x = FoldEnrichment, y = Description)) +
    geom_point(aes(size = Count, colour = neg_log10_padj)) +
    scale_colour_gradient(low = "#DEEBF7", high = "#08519C",
                          name = expression(-log[10](p[adj]))) +
    scale_size_continuous(range = c(3, 10), name = "Gene Count") +
    labs(
      title = paste0("GO ", ont, " — Bubble Plot"),
      x     = "Fold Enrichment",
      y     = NULL
    ) +
    theme_minimal(base_size = 12) +
    theme(plot.title = element_text(face = "bold", size = 14))
  print(p)
}
```

## GO Circle Plot

A custom "GO circle" shows enriched terms as segments of a circle with bar
height proportional to `-log10(padj)` and fill colour encoding the ontology.
This is inspired by the GOplot package style.

```{r go-circle, fig.width=10, fig.height=10}
if (nrow(combined_df) > 0) {
  circle_df <- combined_df %>%
    group_by(Ontology) %>%
    slice_min(p.adjust, n = 8) %>%
    ungroup() %>%
    arrange(Ontology, p.adjust) %>%
    mutate(
      term_label = str_trunc(Description, 40),
      term_id    = row_number()
    )

  n_terms <- nrow(circle_df)

  p <- ggplot(circle_df,
              aes(x = reorder(term_label, term_id),
                  y = neg_log10_padj, fill = Ontology)) +
    geom_col(width = 0.8) +
    coord_polar(start = 0) +
    scale_fill_manual(values = ontology_colours) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed",
               colour = "grey30", linewidth = 0.3) +
    labs(
      title = paste0("GO Circle — Top Enriched Terms by Ontology"),
      x     = NULL,
      y     = expression(-log[10](p[adj]))
    ) +
    theme_minimal(base_size = 11) +
    theme(
      plot.title       = element_text(face = "bold", size = 14, hjust = 0.5),
      axis.text.x      = element_text(size = 8),
      axis.text.y      = element_blank(),
      panel.grid.major = element_line(colour = "grey90", linewidth = 0.2)
    )
  print(p)
}
```

## Waterfall Plot (Ranked Significance)

All significant terms ranked by adjusted p-value, coloured by ontology.  Gives
a quick sense of overall enrichment depth and which ontologies dominate.

```{r waterfall, fig.width=11, fig.height=7}
if (nrow(combined_df) > 0) {
  wf_df <- combined_df %>%
    filter(p.adjust < PVALUE_CUTOFF) %>%
    arrange(p.adjust) %>%
    mutate(Rank = row_number())

  if (nrow(wf_df) > 0) {
    p <- ggplot(wf_df, aes(x = Rank, y = neg_log10_padj,
                            fill = Ontology)) +
      geom_col(width = 0.8) +
      scale_fill_manual(values = ontology_colours) +
      geom_hline(yintercept = -log10(0.05), linetype = "dashed",
                 colour = "grey40") +
      annotate("text", x = max(wf_df$Rank) * 0.9,
               y = -log10(0.05) + 0.3, label = "padj = 0.05",
               size = 3.5, colour = "grey40") +
      labs(
        title = paste0("Waterfall Plot — All Significant GO Terms Ranked"),
        x     = "Rank (ordered by p.adjust)",
        y     = expression(-log[10](p[adj]))
      ) +
      theme_minimal(base_size = 12) +
      theme(plot.title = element_text(face = "bold", size = 14))
    print(p)
  }
}
```

## Fold Enrichment vs. Significance (Volcano-Style)

A scatter plot with fold enrichment on the x-axis and `-log10(padj)` on the
y-axis, sized by gene count.  Analogous to a volcano plot for enrichment.

```{r volcano-enrich, fig.width=11, fig.height=8}
if (nrow(combined_df) > 0) {
  p <- ggplot(combined_df,
              aes(x = log2(FoldEnrichment),
                  y = neg_log10_padj)) +
    geom_point(aes(size = Count, colour = Ontology), alpha = 0.7) +
    scale_colour_manual(values = ontology_colours) +
    scale_size_continuous(range = c(2, 9), name = "Gene Count") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed",
               colour = "grey40") +
    geom_vline(xintercept = 0, linetype = "dotted", colour = "grey60") +
    ggrepel::geom_text_repel(
      data = combined_df %>%
        group_by(Ontology) %>%
        slice_min(p.adjust, n = 3) %>%
        ungroup(),
      aes(label = str_trunc(Description, 35)),
      size = 3, max.overlaps = 15, box.padding = 0.5,
      show.legend = FALSE
    ) +
    labs(
      title = "Enrichment Volcano — Fold Enrichment vs. Significance",
      x     = expression(log[2](Fold~Enrichment)),
      y     = expression(-log[10](p[adj]))
    ) +
    theme_minimal(base_size = 12) +
    theme(plot.title = element_text(face = "bold", size = 14))
  print(p)
}
```

## Dot Matrix (Multi-Ontology Comparison)

Side-by-side comparison of top terms from each ontology in a faceted dot plot.

```{r dot-matrix, fig.width=14, fig.height=10}
if (nrow(combined_df) > 0) {
  facet_df <- combined_df %>%
    group_by(Ontology) %>%
    slice_min(p.adjust, n = 15) %>%
    ungroup() %>%
    mutate(
      Description = str_wrap(Description, width = 40),
      Ontology_label = recode(Ontology,
        "BP" = "Biological Process",
        "MF" = "Molecular Function",
        "CC" = "Cellular Component"
      )
    )

  p <- ggplot(facet_df,
              aes(x = GeneRatioNum, y = reorder(Description, GeneRatioNum))) +
    geom_point(aes(size = Count, colour = neg_log10_padj)) +
    scale_colour_gradient(low = "#FEE8C8", high = "#B30000",
                          name = expression(-log[10](p[adj]))) +
    scale_size_continuous(range = c(2, 8), name = "Gene Count") +
    facet_wrap(~ Ontology_label, scales = "free_y", ncol = 3) +
    labs(
      title = paste0("GO Enrichment Overview — ", SPECIES_NAME),
      x     = "Gene Ratio (k / n)",
      y     = NULL
    ) +
    theme_minimal(base_size = 11) +
    theme(
      plot.title   = element_text(face = "bold", size = 14),
      strip.text   = element_text(face = "bold", size = 12),
      panel.border = element_rect(colour = "grey80", fill = NA)
    )
  print(p)
}
```

## Gene–Term Heatmap (Custom)

A tile plot showing which DE genes map to the top enriched terms, with
hierarchical clustering on both axes.  More control than `heatplot()`.

```{r gene-term-heatmap-custom, fig.width=14, fig.height=9}
for (ont in names(enrich_results)) {
  res <- enrich_results[[ont]]
  if (is.null(res) || nrow(as.data.frame(res)) < 3) next

  res_df <- as.data.frame(res) %>%
    slice_min(p.adjust, n = 20)

  # Build the binary gene × term matrix
  gene_term_list <- res_df %>%
    dplyr::select(ID, Description, geneID) %>%
    mutate(genes = str_split(geneID, "/")) %>%
    unnest(genes) %>%
    dplyr::select(Description, genes)

  mat <- gene_term_list %>%
    mutate(present = 1) %>%
    pivot_wider(names_from = genes, values_from = present, values_fill = 0)

  term_names <- mat$Description
  mat <- as.matrix(mat[, -1])
  rownames(mat) <- term_names

  # Cluster
  if (nrow(mat) >= 2 && ncol(mat) >= 2) {
    row_ord <- hclust(dist(mat, method = "binary"))$order
    col_ord <- hclust(dist(t(mat), method = "binary"))$order
    mat <- mat[row_ord, col_ord]
  }

  heatmap_df <- as.data.frame(mat) %>%
    rownames_to_column("GO_term") %>%
    pivot_longer(-GO_term, names_to = "GeneID", values_to = "Annotated") %>%
    mutate(
      GO_term = factor(GO_term, levels = rownames(mat)),
      GeneID  = factor(GeneID, levels = colnames(mat))
    )

  p <- ggplot(heatmap_df, aes(x = GeneID, y = GO_term, fill = factor(Annotated))) +
    geom_tile(colour = "white", linewidth = 0.3) +
    scale_fill_manual(
      values = c("0" = "grey95", "1" = ontology_colours[ont]),
      labels = c("No", "Yes"), name = "Annotated"
    ) +
    labs(
      title = paste0("GO ", ont, " — Gene × Term Heatmap (clustered)"),
      x     = "Gene ID",
      y     = NULL
    ) +
    theme_minimal(base_size = 11) +
    theme(
      plot.title    = element_text(face = "bold", size = 14),
      axis.text.x   = element_text(angle = 55, hjust = 1, size = 7),
      axis.text.y   = element_text(size = 8),
      legend.position = "bottom"
    )
  print(p)
}
```

## Word Cloud of Enriched Terms

A word cloud built from GO term descriptions weighted by significance.
Requires the `ggwordcloud` package.

```{r wordcloud, fig.width=10, fig.height=8}
if (requireNamespace("ggwordcloud", quietly = TRUE) && nrow(combined_df) > 0) {
  library(ggwordcloud)

  wc_df <- combined_df %>%
    filter(p.adjust < PVALUE_CUTOFF) %>%
    # split multi-word descriptions into tokens
    mutate(words = str_split(Description, "\\s+")) %>%
    unnest(words) %>%
    mutate(words = str_to_lower(words)) %>%
    # remove common stop words
    filter(!words %in% c("of", "to", "in", "by", "the", "a", "an", "and",
                          "or", "via", "from", "with", "into", "for")) %>%
    filter(nchar(words) > 2) %>%
    group_by(words, Ontology) %>%
    summarise(freq = n(), avg_sig = mean(neg_log10_padj), .groups = "drop") %>%
    slice_max(freq, n = 80)

  p <- ggplot(wc_df, aes(label = words, size = freq, colour = Ontology)) +
    ggwordcloud::geom_text_wordcloud(area_corr = TRUE, shape = "circle") +
    scale_size_area(max_size = 16) +
    scale_colour_manual(values = ontology_colours) +
    labs(title = "Word Cloud of Enriched GO Term Descriptions") +
    theme_minimal(base_size = 12) +
    theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 14))
  print(p)
} else if (nrow(combined_df) > 0) {
  message("Install ggwordcloud for word cloud: install.packages('ggwordcloud')")
}
```

## {-}

# Combined Three-Ontology Summary Plots

## Combined Horizontal Bar Plot

Top enriched terms from all three ontologies in a single view, colour-coded.

```{r combined-barplot, fig.width=12, fig.height=9}
if (nrow(combined_df) > 0) {
  bar_df <- combined_df %>%
    group_by(Ontology) %>%
    slice_min(p.adjust, n = 10) %>%
    ungroup() %>%
    mutate(
      Description = str_wrap(Description, width = 45),
      Description = fct_reorder(Description, neg_log10_padj)
    )

  p <- ggplot(bar_df, aes(x = neg_log10_padj, y = Description,
                           fill = Ontology)) +
    geom_col(width = 0.7) +
    scale_fill_manual(values = ontology_colours) +
    geom_vline(xintercept = -log10(0.05), linetype = "dashed",
               colour = "grey40") +
    labs(
      title    = paste0("Top Enriched GO Terms — ", SPECIES_NAME),
      subtitle = "Dashed line = padj = 0.05 threshold",
      x = expression(-log[10](p[adj])),
      y = NULL
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title      = element_text(face = "bold"),
      legend.position = "top"
    )
  print(p)
}
```

## Combined Dot Plot (Publication-Ready)

A compact dot plot designed for journal figures.  Gene ratio on the x-axis,
colour for significance, shape for ontology.

```{r combined-dot-pub, fig.width=12, fig.height=10}
if (nrow(combined_df) > 0) {
  pub_df <- combined_df %>%
    group_by(Ontology) %>%
    slice_min(p.adjust, n = 8) %>%
    ungroup() %>%
    mutate(
      Description = str_wrap(Description, width = 50),
      Description = fct_reorder(Description, GeneRatioNum)
    )

  p <- ggplot(pub_df, aes(x = GeneRatioNum, y = Description)) +
    geom_point(aes(size = Count, colour = neg_log10_padj,
                   shape = Ontology)) +
    scale_colour_viridis_c(option = "inferno", direction = -1,
                           name = expression(-log[10](p[adj]))) +
    scale_shape_manual(values = c("BP" = 16, "MF" = 17, "CC" = 15),
                       name = "Ontology") +
    scale_size_continuous(range = c(3, 10), name = "Gene Count") +
    labs(
      title = paste0("GO Enrichment Summary — ", SPECIES_NAME),
      x     = "Gene Ratio (k / n)",
      y     = NULL
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title      = element_text(face = "bold", size = 14),
      legend.position = "right"
    )
  print(p)
}
```

## Ontology Composition Stacked Bar

Shows how many significant terms belong to each ontology category.

```{r ontology-composition, fig.width=8, fig.height=5}
if (nrow(combined_df) > 0) {
  sig_counts <- combined_df %>%
    filter(p.adjust < PVALUE_CUTOFF) %>%
    count(Ontology, name = "n_terms") %>%
    mutate(
      pct   = n_terms / sum(n_terms) * 100,
      label = paste0(n_terms, " (", round(pct, 1), "%)")
    )

  p <- ggplot(sig_counts, aes(x = "", y = n_terms, fill = Ontology)) +
    geom_col(width = 0.6) +
    geom_text(aes(label = label), position = position_stack(vjust = 0.5),
              size = 4.5, fontface = "bold") +
    scale_fill_manual(values = ontology_colours) +
    coord_flip() +
    labs(
      title = "Distribution of Significant GO Terms by Ontology",
      x     = NULL,
      y     = "Number of enriched terms (padj < 0.05)"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title      = element_text(face = "bold"),
      legend.position = "top",
      axis.text.y     = element_blank()
    )
  print(p)
}
```

# Save All Plots to Files

```{r save-plots, fig.show='hide'}
PLOT_DIR <- file.path(OUT_DIR, "plots")
dir.create(PLOT_DIR, showWarnings = FALSE, recursive = TRUE)

# ── Helper: build the plot inside tryCatch so failures are caught ──
safe_save <- function(filename, plot_expr_fn, w = 10, h = 8) {
  tryCatch({
    p <- plot_expr_fn()            # evaluate inside tryCatch
    ggsave(file.path(PLOT_DIR, filename), plot = p,
           width = w, height = h, dpi = 300, bg = "white")
    message("Saved: ", filename)
  }, error = function(e) {
    message("Skipped ", filename, ": ", e$message)
  })
}

# ── Per-ontology enrichplot plots ──
for (ont in names(enrich_results)) {
  res <- enrich_results[[ont]]
  if (is.null(res) || nrow(as.data.frame(res)) == 0) next

  # Dot plot
  safe_save(paste0("dotplot_", ont, ".png"), function() {
    dotplot(res, showCategory = 20, font.size = 10) +
      ggtitle(paste0("GO ", ont, " — Dot Plot"))
  })

  # Bar plot
  safe_save(paste0("barplot_", ont, ".png"), function() {
    enrich_barplot(res, showCategory = 20, font.size = 10) +
      ggtitle(paste0("GO ", ont, " — Bar Plot"))
  })

  # Enrichment map
  if (nrow(as.data.frame(res)) >= 2) {
    res_pw <- pairwise_termsim(res)

    safe_save(paste0("emap_", ont, ".png"), function() {
      emapplot(res_pw, showCategory = 30) +
        ggtitle(paste0("GO ", ont, " — Enrichment Map"))
    }, w = 12, h = 10)

    # Clustered bar plot (replaces treeplot)
    if (nrow(as.data.frame(res)) >= 3) {
      safe_save(paste0("clustered_bar_", ont, ".png"), function() {
        res_df <- as.data.frame(res) %>% head(25)
        gene_lists <- strsplit(res_df$geneID, "/")
        all_genes  <- unique(unlist(gene_lists))
        mat <- matrix(0, nrow = nrow(res_df), ncol = length(all_genes),
                      dimnames = list(res_df$Description, all_genes))
        for (i in seq_along(gene_lists)) mat[i, gene_lists[[i]]] <- 1
        hc <- hclust(dist(mat, method = "binary"), method = "ward.D2")
        res_df$Description <- factor(res_df$Description,
                                      levels = res_df$Description[hc$order])
        ggplot(res_df, aes(x = -log10(p.adjust), y = Description,
                           fill = -log10(p.adjust))) +
          geom_col(width = 0.7) +
          scale_fill_gradient(low = "#FDD49E", high = "#B30000") +
          labs(title = paste0("GO ", ont, " — Clustered Enrichment"),
               x = expression(-log[10](p[adj])), y = NULL) +
          theme_minimal(base_size = 11) +
          theme(plot.title = element_text(face = "bold"))
      }, w = 14, h = 10)
    }

    # Heatmap
    safe_save(paste0("heatplot_", ont, ".png"), function() {
      heatplot(res, showCategory = 20) +
        ggtitle(paste0("GO ", ont, " — Heatmap"))
    }, w = 14, h = 9)

    # Category-gene network
    safe_save(paste0("cnetplot_", ont, ".png"), function() {
      cnetplot(res, showCategory = 10) +
        ggtitle(paste0("GO ", ont, " — Category–Gene Network"))
    }, w = 13, h = 10)
  }
}

# ── Combined plots ──
if (nrow(combined_df) > 0) {
  # Combined bar
  safe_save("combined_barplot.png", function() {
    bar_df <- combined_df %>%
      group_by(Ontology) %>% slice_min(p.adjust, n = 10) %>% ungroup() %>%
      mutate(Description = str_wrap(Description, 45),
             Description = fct_reorder(Description, neg_log10_padj))
    ggplot(bar_df, aes(x = neg_log10_padj, y = Description, fill = Ontology)) +
      geom_col(width = 0.7) +
      scale_fill_manual(values = ontology_colours) +
      geom_vline(xintercept = -log10(0.05), linetype = "dashed", colour = "grey40") +
      labs(title = paste0("Top Enriched GO Terms — ", SPECIES_NAME),
           x = expression(-log[10](p[adj])), y = NULL) +
      theme_minimal(base_size = 13) +
      theme(plot.title = element_text(face = "bold"), legend.position = "top")
  }, w = 12, h = 9)

  # Volcano-style
  safe_save("enrichment_volcano.png", function() {
    ggplot(combined_df, aes(x = log2(FoldEnrichment), y = neg_log10_padj)) +
      geom_point(aes(size = Count, colour = Ontology), alpha = 0.7) +
      scale_colour_manual(values = ontology_colours) +
      scale_size_continuous(range = c(2, 9), name = "Gene Count") +
      geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey40") +
      labs(title = "Enrichment Volcano", x = expression(log[2](Fold~Enrichment)),
           y = expression(-log[10](p[adj]))) +
      theme_minimal(base_size = 12) +
      theme(plot.title = element_text(face = "bold", size = 14))
  }, w = 11, h = 8)

  # Faceted dot matrix
  safe_save("dot_matrix_faceted.png", function() {
    facet_df <- combined_df %>%
      group_by(Ontology) %>% slice_min(p.adjust, n = 15) %>% ungroup() %>%
      mutate(Description = str_wrap(Description, 40),
             Ontology_label = recode(Ontology,
               "BP" = "Biological Process", "MF" = "Molecular Function",
               "CC" = "Cellular Component"))
    ggplot(facet_df, aes(x = GeneRatioNum,
                         y = reorder(Description, GeneRatioNum))) +
      geom_point(aes(size = Count, colour = neg_log10_padj)) +
      scale_colour_gradient(low = "#FEE8C8", high = "#B30000",
                            name = expression(-log[10](p[adj]))) +
      scale_size_continuous(range = c(2, 8), name = "Gene Count") +
      facet_wrap(~ Ontology_label, scales = "free_y", ncol = 3) +
      labs(title = paste0("GO Enrichment Overview — ", SPECIES_NAME),
           x = "Gene Ratio", y = NULL) +
      theme_minimal(base_size = 11) +
      theme(plot.title = element_text(face = "bold", size = 14),
            strip.text = element_text(face = "bold", size = 12),
            panel.border = element_rect(colour = "grey80", fill = NA))
  }, w = 14, h = 10)
}

cat("\n--- All plots saved to:", PLOT_DIR, "---\n")
list.files(PLOT_DIR, pattern = "\\.png$")
```

# Export Results

```{r export}
# ── Save per-ontology TSV files ──
imap(enrich_results, function(res, ont) {
  if (!is.null(res) && nrow(as.data.frame(res)) > 0) {
    out_file <- file.path(OUT_DIR, paste0("GO_enrichment_", ont, ".tsv"))
    write.table(
      as.data.frame(res),
      file      = out_file,
      sep       = "\t",
      row.names = FALSE,
      quote     = FALSE
    )
    message("Saved: ", out_file)
  }
})

# ── Save combined results ──
if (nrow(combined_df) > 0) {
  combined_out <- file.path(OUT_DIR, "GO_enrichment_combined.tsv")
  write.table(combined_df, file = combined_out, sep = "\t",
              row.names = FALSE, quote = FALSE)
  message("Saved: ", combined_out)
}
```

# Manual Fisher's Exact Test (for Reference)

For transparency, here is the manual implementation of the same over-
representation test that `enricher()` performs internally.  This can be useful
for debugging or for understanding the statistics.

```{r manual-fisher}
# Pick one ontology for demonstration
demo_t2g <- t2g_list[["BP"]]
demo_t2n <- t2n_list[["BP"]]

# IMPORTANT: restrict N and n to genes that appear in the BP TERM2GENE,
# matching what enricher() does internally.  Using the full background
# (all ontologies) would dilute N and n with genes that have no BP
# annotation, making the test slightly more conservative than enricher().
bp_universe <- intersect(bg_chars, unique(demo_t2g$gene))
bp_de       <- intersect(de_chars, unique(demo_t2g$gene))
N <- length(bp_universe)
n <- length(bp_de)

cat(sprintf(
  "Manual Fisher (BP): N = %d background genes, n = %d DE genes (BP-annotated only)\n",
  N, n
))

manual_results <- demo_t2g %>%
  group_by(term) %>%
  summarise(
    K = n_distinct(gene),                                    # genes in background with this term
    k = sum(gene %in% de_chars),                             # DE genes with this term
    .groups = "drop"
  ) %>%
  filter(k >= 2, K >= MIN_GENE_SET, K <= MAX_GENE_SET) %>%  # minimum overlap & set size
  rowwise() %>%
  mutate(
    pvalue = fisher.test(
      matrix(c(k, K - k, n - k, N - K - n + k), nrow = 2),
      alternative = "greater"
    )$p.value,
    fold_enrichment = (k / n) / (K / N),
    GeneRatio = paste0(k, "/", n),
    BgRatio   = paste0(K, "/", N)
  ) %>%
  ungroup() %>%
  mutate(padj = p.adjust(pvalue, method = "BH")) %>%
  left_join(demo_t2n, by = "term") %>%
  arrange(padj)

manual_results %>%
  filter(padj < PVALUE_CUTOFF) %>%
  dplyr::select(term, name, k, K, GeneRatio, BgRatio, fold_enrichment, pvalue, padj) %>%
  mutate(across(c(pvalue, padj), ~ signif(.x, 3)),
         fold_enrichment = round(fold_enrichment, 2)) %>%
  datatable(
    rownames = FALSE,
    caption  = "Manual Fisher's exact test results (BP) — should match enricher() output",
    options  = list(pageLength = 15, scrollX = TRUE)
  )
```

# Tips for Your Own Analysis

## Choosing the Right Background

The **background (universe)** profoundly affects your results:
  
- **Best practice**: use all genes that were *expressed and testable* in your
  experiment (e.g., all genes that passed DESeq2's independent filtering).
- **Acceptable fallback**: all genes in `gene2go` for your organism (what this
  pipeline uses by default).
- **Avoid**: using the entire genome (inflates significance) or no universe at
  all (enricher defaults to all genes in TERM2GENE, which is correct only if
  every gene was testable).

## Supplementing Sparse Annotations

For non-model organisms, `gene2go` annotations are often limited to IEA
(electronic inference).  To improve coverage:

1. **eggNOG-mapper** (<http://eggnog-mapper.embl.de/>): upload your protein
   FASTA; returns GO, KEGG, COG annotations via ortholog mapping.
2. **InterProScan**: domain-based GO assignment for every protein.
3. **Trinotate** (if using Trinity for *de novo* assembly): wraps BLAST,
   HMMER, and GO assignment into one pipeline.

After obtaining additional annotations, merge them into `TERM2GENE` before
running `enricher()`.

## GO Hierarchy (DAG Propagation) — topGO Analysis

A gene annotated with GO:0006954 (inflammatory response) should also be counted
under its ancestor GO:0006952 (defense response).  `enricher()` does **not**
propagate annotations up the GO directed acyclic graph (DAG).

`topGO` explicitly traverses the hierarchy and offers algorithms that
decorrelate parent–child redundancy:

- **classic**: standard Fisher's exact test (no hierarchy awareness — baseline).
- **weight01** (recommended): a hybrid of the *weight* and *elim* algorithms
  that down-weights broad parent terms when their children are already
  significant, yielding more specific, interpretable results.
- **elim**: removes genes from parent terms once a child term is significant.

```{r topgo-setup}
if (!requireNamespace("topGO", quietly = TRUE)) {
  message("topGO not installed — skipping DAG-aware enrichment.\n",
          "Install with: BiocManager::install('topGO')")
  RUN_TOPGO <- FALSE
} else {
  library(topGO)
  RUN_TOPGO <- TRUE
}
```

### Run topGO for All Three Ontologies {.tabset}

```{r topgo-run}
if (RUN_TOPGO) {
  # ── Build gene-to-GO mapping (named list) ──
  gene2GO_list <- split(gene2go_sp$GO_ID, as.character(gene2go_sp$GeneID))

  # ── Named factor: 1 = DE, 0 = not DE ──
  all_genes <- factor(as.integer(bg_chars %in% de_chars))
  names(all_genes) <- bg_chars

  # ── Run for each ontology ──
  ontologies <- c("BP", "MF", "CC")
  topgo_results <- list()

  for (ont in ontologies) {
    message("Running topGO for ", ont, " ...")

    # Build topGOdata object
    GOdata <- tryCatch(
      new("topGOdata",
          ontology  = ont,
          allGenes  = all_genes,
          annot     = annFUN.gene2GO,
          gene2GO   = gene2GO_list,
          nodeSize  = MIN_GENE_SET),
      error = function(e) {
        message("  topGOdata failed for ", ont, ": ", e$message)
        NULL
      }
    )
    if (is.null(GOdata)) next

    # Classic Fisher (baseline)
    res_classic  <- runTest(GOdata, algorithm = "classic",
                             statistic = "fisher")
    # Weight01 (DAG-aware, recommended)
    res_weight01 <- runTest(GOdata, algorithm = "weight01",
                             statistic = "fisher")
    # Elim (alternative DAG-aware)
    res_elim     <- runTest(GOdata, algorithm = "elim",
                             statistic = "fisher")

    # Combine into a summary table
    n_terms <- min(50, length(usedGO(GOdata)))
    allRes <- GenTable(
      GOdata,
      classic  = res_classic,
      weight01 = res_weight01,
      elim     = res_elim,
      orderBy  = "weight01",
      ranksOf  = "classic",
      topNodes = n_terms
    )

    # Clean up p-value columns (GenTable returns character with "< 1e-30" etc.)
    allRes <- allRes %>%
      mutate(
        across(c(classic, weight01, elim), ~ {
          x <- gsub("< ", "", .x)
          as.numeric(x)
        }),
        Ontology = ont,
        # BH correction on classic only — weight01/elim account for GO DAG
        # structure and their p-values should be used directly without further
        # BH adjustment (Alexa & Rahnenführer, 2009)
        classic_padj = p.adjust(classic, method = "BH")
      )

    topgo_results[[ont]] <- allRes
    message("  ", ont, ": ", sum(allRes$weight01 < PVALUE_CUTOFF, na.rm = TRUE),
            " terms with weight01 p < ", PVALUE_CUTOFF)
  }
}
```

#### BP — topGO Results

```{r topgo-bp-table}
if (RUN_TOPGO && !is.null(topgo_results[["BP"]])) {
  topgo_results[["BP"]] %>%
    mutate(across(c(classic, weight01, elim, classic_padj),
                  ~ signif(.x, 3))) %>%
    datatable(
      rownames = FALSE,
      caption  = "topGO Biological Process — weight01 + classic + elim",
      options  = list(pageLength = 20, scrollX = TRUE)
    )
} else {
  cat("topGO BP results not available.\n")
}
```

#### MF — topGO Results

```{r topgo-mf-table}
if (RUN_TOPGO && !is.null(topgo_results[["MF"]])) {
  topgo_results[["MF"]] %>%
    mutate(across(c(classic, weight01, elim, classic_padj),
                  ~ signif(.x, 3))) %>%
    datatable(
      rownames = FALSE,
      caption  = "topGO Molecular Function — weight01 + classic + elim",
      options  = list(pageLength = 20, scrollX = TRUE)
    )
} else {
  cat("topGO MF results not available.\n")
}
```

#### CC — topGO Results

```{r topgo-cc-table}
if (RUN_TOPGO && !is.null(topgo_results[["CC"]])) {
  topgo_results[["CC"]] %>%
    mutate(across(c(classic, weight01, elim, classic_padj),
                  ~ signif(.x, 3))) %>%
    datatable(
      rownames = FALSE,
      caption  = "topGO Cellular Component — weight01 + classic + elim",
      options  = list(pageLength = 20, scrollX = TRUE)
    )
} else {
  cat("topGO CC results not available.\n")
}
```

### {-}

### topGO Visualisations {.tabset}

#### Algorithm Comparison (Classic vs. Weight01)

Scatter plot comparing p-values from the classic (no hierarchy) and weight01
(DAG-aware) algorithms.  Points below the diagonal are terms that become *more*
significant after hierarchy correction; points above become *less* significant
(broad parent terms down-weighted).

```{r topgo-scatter, fig.width=10, fig.height=8}
if (RUN_TOPGO && length(topgo_results) > 0) {
  topgo_combined <- bind_rows(topgo_results)

  p <- ggplot(topgo_combined,
              aes(x = -log10(classic), y = -log10(weight01))) +
    geom_point(aes(colour = Ontology, size = Significant), alpha = 0.6) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                colour = "grey40") +
    scale_colour_manual(values = ontology_colours) +
    scale_size_continuous(range = c(1.5, 7), name = "DE Genes\nin Term") +
    ggrepel::geom_text_repel(
      data = topgo_combined %>%
        group_by(Ontology) %>%
        slice_min(weight01, n = 5) %>%
        ungroup(),
      aes(label = str_trunc(Term, 35)),
      size = 3, max.overlaps = 15, box.padding = 0.5,
      show.legend = FALSE
    ) +
    labs(
      title = "topGO: Classic vs. Weight01 Algorithm",
      subtitle = "Points below diagonal = more significant after DAG correction",
      x = expression(-log[10](p[classic])),
      y = expression(-log[10](p[weight01]))
    ) +
    theme_minimal(base_size = 12) +
    theme(plot.title = element_text(face = "bold", size = 14))
  print(p)
}
```

#### topGO Bar Plot (Weight01)

Top enriched terms from topGO's weight01 algorithm across all ontologies.

```{r topgo-barplot, fig.width=12, fig.height=9}
if (RUN_TOPGO && length(topgo_results) > 0) {
  topgo_bar <- topgo_combined %>%
    filter(weight01 < PVALUE_CUTOFF) %>%
    group_by(Ontology) %>%
    slice_min(weight01, n = 10) %>%
    ungroup() %>%
    mutate(
      Term = str_wrap(Term, width = 45),
      Term = fct_reorder(Term, -log10(weight01))
    )

  if (nrow(topgo_bar) > 0) {
    p <- ggplot(topgo_bar, aes(x = -log10(weight01), y = Term,
                                fill = Ontology)) +
      geom_col(width = 0.7) +
      scale_fill_manual(values = ontology_colours) +
      geom_vline(xintercept = -log10(0.05), linetype = "dashed",
                 colour = "grey40") +
      labs(
        title = "topGO Weight01 — Top Enriched GO Terms",
        x     = expression(-log[10](p[weight01])),
        y     = NULL
      ) +
      theme_minimal(base_size = 12) +
      theme(plot.title = element_text(face = "bold", size = 14),
            legend.position = "top")
    print(p)
  } else {
    cat("No terms with weight01 p <", PVALUE_CUTOFF, "\n")
  }
}
```

#### {-}

### enricher() vs. topGO Comparison {.tabset}

Which terms appear in both methods vs. only one?  Terms found by both are high
confidence; terms unique to topGO are typically more specific (hierarchy-aware),
while terms unique to enricher may be broad parent terms inflated by
unadjusted annotation propagation.

> **Note on thresholds:** enricher results are filtered by BH-adjusted p-value
> (`p.adjust < 0.05`), while topGO weight01 results use the raw algorithm
> p-value (`weight01 < 0.05`).  This follows the recommended practice for each
> method — weight01 p-values already account for DAG structure and should not
> be BH-corrected further (Alexa & Rahnenführer, 2009).  Consequently, the
> comparison is not strictly apples-to-apples: a term appearing as "topGO only"
> may partly reflect the less stringent threshold rather than a biological
> difference.

```{r comparison-table}
if (RUN_TOPGO && length(topgo_results) > 0 && nrow(combined_df) > 0) {
  comparison_list <- list()

  for (ont in names(enrich_results)) {
    # enricher significant terms
    enricher_sig <- combined_df %>%
      filter(Ontology == ont, p.adjust < PVALUE_CUTOFF) %>%
      dplyr::select(GO_ID = ID, Description,
                    enricher_padj = p.adjust, enricher_count = Count)

    # topGO significant terms (weight01)
    if (!is.null(topgo_results[[ont]])) {
      topgo_sig <- topgo_results[[ont]] %>%
        filter(weight01 < PVALUE_CUTOFF) %>%
        dplyr::select(GO_ID = GO.ID, Term,
                      topGO_weight01 = weight01, topGO_annotated = Annotated,
                      topGO_significant = Significant)

      comp <- full_join(enricher_sig, topgo_sig, by = "GO_ID") %>%
        mutate(
          Found_in = case_when(
            !is.na(enricher_padj) & !is.na(topGO_weight01) ~ "Both",
            !is.na(enricher_padj)                           ~ "enricher only",
            TRUE                                             ~ "topGO only"
          ),
          Name = coalesce(Description, Term),
          Ontology = ont
        ) %>%
        dplyr::select(Ontology, GO_ID, Name, Found_in,
                      enricher_padj, topGO_weight01)

      comparison_list[[ont]] <- comp
    }
  }

  comparison_df <- bind_rows(comparison_list)
}
```

#### Comparison Table

```{r comparison-dt}
if (exists("comparison_df") && nrow(comparison_df) > 0) {
  comparison_df %>%
    mutate(across(c(enricher_padj, topGO_weight01), ~ signif(.x, 3))) %>%
    arrange(Found_in, Ontology) %>%
    datatable(
      rownames = FALSE,
      caption  = "enricher() vs. topGO weight01 — term overlap",
      filter   = "top",
      options  = list(pageLength = 25, scrollX = TRUE)
    )
} else {
  cat("Comparison not available.\n")
}
```

#### Overlap Summary

```{r comparison-summary, fig.width=10, fig.height=6}
if (exists("comparison_df") && nrow(comparison_df) > 0) {
  overlap_counts <- comparison_df %>%
    count(Ontology, Found_in, name = "n_terms")

  p <- ggplot(overlap_counts,
              aes(x = Ontology, y = n_terms, fill = Found_in)) +
    geom_col(position = "dodge", width = 0.7) +
    geom_text(aes(label = n_terms),
              position = position_dodge(width = 0.7), vjust = -0.3,
              size = 4, fontface = "bold") +
    scale_fill_manual(
      values = c("Both"          = "#2C7BB6",
                 "enricher only" = "#D7191C",
                 "topGO only"    = "#1A9641"),
      name = "Found in"
    ) +
    labs(
      title = "enricher() vs. topGO — Term Overlap by Ontology",
      subtitle = "'Both' = high-confidence terms; 'topGO only' = hierarchy-refined specifics",
      x = NULL,
      y = "Number of significant terms"
    ) +
    theme_minimal(base_size = 13) +
    theme(plot.title = element_text(face = "bold", size = 14),
          legend.position = "top")
  print(p)
}
```

#### Significance Concordance

For terms found by both methods, how well do the p-values agree?

```{r concordance-scatter, fig.width=10, fig.height=8}
if (exists("comparison_df") && nrow(comparison_df) > 0) {
  both_df <- comparison_df %>%
    filter(Found_in == "Both") %>%
    filter(!is.na(enricher_padj), !is.na(topGO_weight01))

  if (nrow(both_df) >= 3) {
    p <- ggplot(both_df,
                aes(x = -log10(enricher_padj),
                    y = -log10(topGO_weight01))) +
      geom_point(aes(colour = Ontology), size = 3, alpha = 0.7) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                  colour = "grey40") +
      scale_colour_manual(values = ontology_colours) +
      ggrepel::geom_text_repel(
        data = both_df %>% slice_min(enricher_padj, n = 8),
        aes(label = str_trunc(Name, 35)),
        size = 3, max.overlaps = 12, show.legend = FALSE
      ) +
      labs(
        title = "Significance Concordance — enricher() vs. topGO",
        subtitle = "Diagonal shown for reference (axes use different correction methods)",
        x = expression(-log[10](enricher~p[adj]~(BH))),
        y = expression(-log[10](topGO~weight01~p~(raw)))
      ) +
      theme_minimal(base_size = 12) +
      theme(plot.title = element_text(face = "bold", size = 14))
    print(p)
  } else {
    cat("Fewer than 3 terms found by both methods — skipping concordance plot.\n")
  }
}
```

#### {-}

### Export topGO Results

```{r topgo-export}
if (RUN_TOPGO && length(topgo_results) > 0) {
  # Per-ontology topGO results
  for (ont in names(topgo_results)) {
    out_file <- file.path(OUT_DIR, paste0("topGO_", ont, ".tsv"))
    write.table(topgo_results[[ont]], file = out_file, sep = "\t",
                row.names = FALSE, quote = FALSE)
    message("Saved: ", out_file)
  }

  # Comparison table
  if (exists("comparison_df") && nrow(comparison_df) > 0) {
    comp_file <- file.path(OUT_DIR, "enricher_vs_topGO_comparison.tsv")
    write.table(comparison_df, file = comp_file, sep = "\t",
                row.names = FALSE, quote = FALSE)
    message("Saved: ", comp_file)
  }
}
```

# Session Info

```{r session-info}
sessionInfo()
```
